---
title: "Reference Sequence Similarity QC"
author: "aa-tRNA-seq Pipeline"
format:
  html:
    code-fold: true
    toc: true
params:
  similarity_matrix: "summary/qc/reference_similarity.tsv"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Overview

This report visualizes pairwise sequence similarity across all sequences in the
reference FASTA file. High similarity between sequences may indicate potential
cross-mapping issues during alignment.

## Similarity Matrix Heatmap

```{r load-data}
library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(tidyr)
library(stringr)

# Read similarity matrix
mat <- as.matrix(read.table(params$similarity_matrix, header = TRUE,
                             row.names = 1, sep = "\t", check.names = FALSE))
```

```{r parse-names}
# Parse tRNA names to extract amino acid and anticodon
# Expected format: tRNA-Ala-AGC-1-1 or similar
parse_trna_name <- function(name) {
  parts <- str_split(name, "-")[[1]]

  # Handle different naming conventions
  if (length(parts) >= 3 && parts[1] == "tRNA") {
    aa <- parts[2]
    anticodon <- parts[3]
  } else if (length(parts) >= 2) {
    # Fallback for other formats
    aa <- parts[1]
    anticodon <- if (length(parts) >= 2) parts[2] else "unknown"
  } else {
    aa <- "unknown"
    anticodon <- "unknown"
  }

  list(aa = aa, anticodon = anticodon)
}

# Parse all sequence names
parsed <- lapply(rownames(mat), parse_trna_name)
aa_labels <- sapply(parsed, function(x) x$aa)
anticodon_labels <- sapply(parsed, function(x) x$anticodon)
```

```{r color-setup}
# Create color palette for amino acids
unique_aa <- unique(aa_labels)
n_aa <- length(unique_aa)

# Use a qualitative palette for amino acids
if (n_aa <= 20) {
  aa_colors <- setNames(
    rainbow(n_aa, s = 0.7, v = 0.8),
    unique_aa
  )
} else {
  aa_colors <- setNames(
    colorRampPalette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
                       "#FF7F00", "#FFFF33", "#A65628", "#F781BF"))(n_aa),
    unique_aa
  )
}

# Color scale for similarity values
# Low similarity (50%) = white/blue, High similarity (100%) = red
col_fun <- colorRamp2(c(50, 75, 90, 100), c("white", "lightyellow", "orange", "red"))
```

```{r heatmap, fig.width=12, fig.height=10}
# Create row annotation for amino acid
row_anno <- rowAnnotation(
  AA = aa_labels,
  col = list(AA = aa_colors),
  show_annotation_name = TRUE,
  annotation_name_side = "top"
)

# Create column annotation for amino acid
col_anno <- HeatmapAnnotation(
  AA = aa_labels,
  col = list(AA = aa_colors),
  show_annotation_name = FALSE
)

# Draw heatmap with clustering
ht <- Heatmap(
  mat,
  name = "% Identity",
  col = col_fun,
  left_annotation = row_anno,
  top_annotation = col_anno,
  show_row_names = nrow(mat) <= 50,
  show_column_names = nrow(mat) <= 50,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 6),
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",
  row_title = "Reference Sequences",
  column_title = "Reference Sequence Similarity"
)

draw(ht)
```

## Highly Similar Sequence Pairs

Sequence pairs with >90% identity may be at risk for cross-mapping:

```{r high-similarity-pairs}
# Find pairs with high similarity (excluding diagonal)
high_sim_threshold <- 90

# Get upper triangle indices (excluding diagonal).
# Indexed rather than looped: a nested loop over the matrix is O(n^2)
# interpreted iterations and does not finish on a large reference.
hits <- which(mat >= high_sim_threshold & upper.tri(mat), arr.ind = TRUE)

pairs <- data.frame(
  Seq1 = rownames(mat)[hits[, "row"]],
  Seq2 = colnames(mat)[hits[, "col"]],
  Percent_Identity = round(mat[hits], 2)
)

# Sort by similarity (descending)
pairs <- pairs %>% arrange(desc(Percent_Identity))

if (nrow(pairs) > 0) {
  knitr::kable(head(pairs, 50),
               caption = paste0("Top sequence pairs with >= ", high_sim_threshold, "% identity"))
} else {
  cat(paste0("No sequence pairs found with >= ", high_sim_threshold, "% identity."))
}
```

## Summary Statistics

```{r summary-stats}
# Get upper triangle values (excluding diagonal)
upper_tri <- mat[upper.tri(mat)]

cat("Similarity Distribution (excluding self-comparisons):\n")
cat(sprintf("  Min: %.2f%%\n", min(upper_tri)))
cat(sprintf("  Median: %.2f%%\n", median(upper_tri)))
cat(sprintf("  Mean: %.2f%%\n", mean(upper_tri)))
cat(sprintf("  Max: %.2f%%\n", max(upper_tri)))
cat(sprintf("  Pairs >= 90%%: %d\n", sum(upper_tri >= 90)))
cat(sprintf("  Pairs >= 95%%: %d\n", sum(upper_tri >= 95)))
cat(sprintf("  Total pairs: %d\n", length(upper_tri)))
```

```{r similarity-histogram, fig.width=8, fig.height=4}
hist(upper_tri,
     breaks = 50,
     main = "Distribution of Pairwise Sequence Similarities",
     xlab = "Percent Identity",
     col = "steelblue",
     border = "white")
abline(v = 90, col = "red", lty = 2, lwd = 2)
legend("topleft", legend = "90% threshold", col = "red", lty = 2, lwd = 2)
```

## Interpretation

- **Isodecoders** (same anticodon, different genomic copy) are expected to show
  high similarity, as they encode functionally similar tRNAs.

- **Charged/uncharged pairs** of the same tRNA will differ primarily in the 3'
  adapter sequence, showing high similarity in the tRNA body.

- **Cross-family similarity** may indicate alignment ambiguity. Reads mapping
  to highly similar sequences should be interpreted with caution.

- For sequences with >95% identity, consider examining the alignment positions
  of reads mapping to these sequences.
