---
title: "aa-tRNA-seq Combined QC Report"
author: "Pipeline"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
params:
  config_file: "config/config.yml"
  ml_threshold: 200
---

## Pipeline Overview

```{mermaid}
flowchart LR
  A[POD5 files] --> B[Merge POD5s]
  B --> C[Rebasecall<br>Dorado]
  C --> D[FASTQ]
  D --> E[Align<br>BWA MEM]
  E --> F[Classify Charging<br>Remora]
  F --> G[Transfer Tags]
  G --> H[Add Adapter Tags]
  H --> I[Summary Tables<br>& QC Metrics]
```

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(paletteer)
library(fs)
library(patchwork)
library(scales)
library(yaml)
library(cowplot)
library(ComplexHeatmap)
library(circlize)

# Read pipeline config
config <- read_yaml(params$config_file)
output_dir <- fs::path_abs(config$output_directory, start = dirname(params$config_file))
ml_threshold <- params$ml_threshold

# Parse samples from config
# Supports both YAML format (with runs) and TSV format
if (!is.null(config$samples))
{
  samples_path <- fs::path_abs(config$samples, start = dirname(params$config_file))
  if (fs::path_ext(samples_path) %in% c("yml", "yaml")) {
    samples_config <- read_yaml(samples_path)
    sample_metadata <- samples_config$runs |>
      purrr::map_dfr(function(run) {
        tibble(
          sample = names(run$samples),
          barcode = unlist(run$samples),
          run_path = run$path,
          barcode_kit = run$barcode_kit %||% NA_character_
        )
      })
    sample_list <- sample_metadata$sample
  } else {
    # TSV format: first column is sample name, second is run directory
    sample_metadata <- read_tsv(samples_path, col_names = c("sample", "run_path"),
                                show_col_types = FALSE) |>
      group_by(sample) |>
      summarise(run_path = paste(run_path, collapse = "; "), .groups = "drop")
    sample_list <- sample_metadata$sample
  }
} else {
 stop("No samples defined in config file")
}

# Apply sample ordering from config if specified
if (!is.null(config$report$sample_order)) {
  specified_order <- config$report$sample_order
  # Put specified samples first (in order), then any remaining samples
  sample_list <- c(
    intersect(specified_order, sample_list),
    setdiff(sample_list, specified_order)
  )
}

# Extract isodecoder from gtrnadb-format tRNA names
# Pattern: Genus_species_tRNA-Aaa-NNN-X-Y → Aaa-NNN (handles iMet with 4 chars)
# Returns list with isodecoder name and whether it matched the tRNA pattern
extract_isodecoder <- function(trna_name) {
  isodecoder <- str_extract(trna_name, "(?<=tRNA-)[A-Za-z]{3,4}-[ACGT]{3}")
  if_else(is.na(isodecoder), trna_name, isodecoder)
}

# Check if name matches standard tRNA isodecoder pattern
is_standard_trna <- function(trna_name) {
 str_detect(trna_name, "tRNA-[A-Za-z]{3,4}-[ACGT]{3}")
}
```

This report summarizes QC metrics for **`r length(sample_list)` samples** from the aa-tRNA-seq pipeline.

### Samples

```{r}
#| label: sample-table

if ("barcode" %in% colnames(sample_metadata)) {
  sample_metadata |>
    dplyr::select(Sample = sample, `Run Path` = run_path,
                  Barcode = barcode, `Barcode Kit` = barcode_kit) |>
    gt()
} else {
  sample_metadata |>
    dplyr::select(Sample = sample, `Run Path` = run_path) |>
    gt()
}
```

### Pipeline Configuration

```{r}
#| label: config-table

config_params <- tibble(
  Parameter = c(
    "Base Calling Model",
    "Reference FASTA",
    "Dorado Version",
    "Charging ML Threshold",
    "BWA Options",
    "BAM Filter",
    "5' Adapter",
    "3' Adapter"
  ),
  Value = c(
    config$base_calling_model %||% config$dorado_model %||% "N/A",
    config$fasta %||% "N/A",
    as.character(config$dorado_version %||% "N/A"),
    as.character(ml_threshold),
    trimws(config$opts$bwa %||% "N/A"),
    trimws(config$opts$bam_filter %||% "N/A"),
    config$adapters$five_prime %||% "N/A",
    if (is.list(config$adapters$three_prime)) {
      paste(sapply(config$adapters$three_prime, function(x) paste0(x$name, ": ", x$seq)), collapse = "; ")
    } else {
      config$adapters$three_prime %||% "N/A"
    }
  )
)

gt(config_params)
```

```{r}
#| label: load-all-data
#| include: false

# Load all data upfront
align_stats_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".align_stats.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

charging_prob_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".charging_prob.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

charging_cpm_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".charging.cpm.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

bcerror_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".bcerror.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

# Convert sample to factor with correct ordering for all datasets
align_stats_all <- align_stats_all |> mutate(sample = factor(sample, levels = sample_list))
charging_prob_all <- charging_prob_all |> mutate(sample = factor(sample, levels = sample_list))
charging_cpm_all <- charging_cpm_all |> mutate(sample = factor(sample, levels = sample_list))
bcerror_all <- bcerror_all |> mutate(sample = factor(sample, levels = sample_list))

# Load reference similarity matrix
similarity_matrix_path <- fs::path(output_dir, "summary", "qc", "reference_similarity.tsv")
similarity_matrix <- NULL
if (fs::file_exists(similarity_matrix_path)) {
  similarity_df <- read_tsv(similarity_matrix_path, show_col_types = FALSE)
  similarity_matrix <- as.matrix(similarity_df[, -1])
  rownames(similarity_matrix) <- similarity_df[[1]]
}
```

## Alignment Summary

### Read Progression Through Pipeline

::: {.panel-tabset}

```{r}
#| label: alignment-summary-tabs
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_align <- align_stats_all |>
    filter(sample == s) |>
    dplyr::select(
      Stage = info,
      `Total Reads` = n_reads,
      `Mapped Reads` = mapped_reads,
      `% Mapped` = pct_mapped,
      `Mean Length` = mean_length,
      `Median Length` = median_length,
      `Mean Quality` = mean_base_quality,
      `Mean MAPQ` = mean_MAPQ
    ) |>
    mutate(
      Stage = factor(Stage, levels = c("unmapped", "aligned", "classified")),
      Stage = recode(Stage,
        "unmapped" = "Rebasecalled",
        "aligned" = "Aligned",
        "classified" = "Classified"
      )
    )

  tbl <- gt(tbl_align) |>
    fmt_number(columns = c(`Mean Length`, `Mean Quality`, `Mean MAPQ`), decimals = 1) |>
    fmt_number(columns = `% Mapped`, decimals = 1)

  cat(as_raw_html(tbl))
  cat("\n\n")
}
```

:::

### Alignment Funnel

```{r}
#| label: fig-alignment-funnel
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Read counts at each processing stage showing retention through the pipeline."

tbl_funnel <- align_stats_all |>
  dplyr::select(sample, info, n_reads) |>
  mutate(
    stage = factor(info, levels = c("unmapped", "aligned", "classified")),
    stage = recode(stage,
      "unmapped" = "Rebasecalled",
      "aligned" = "Aligned",
      "classified" = "Classified"
    )
  )

ggplot(tbl_funnel, aes(x = stage, y = n_reads, fill = stage)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = comma(n_reads)), vjust = -0.5, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Processing Stage", y = "Number of Reads") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Charging Classification

### Charging Likelihood Distribution

The Remora model assigns a charging likelihood score (0-255) to each read. Reads with scores >= `r ml_threshold` are classified as charged.

```{r}
#| label: fig-charging-density
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Distribution of Remora charging likelihood scores. Dashed line indicates classification threshold."

tbl_density <- charging_prob_all |>
  mutate(
    charging_status = if_else(charging_likelihood >= ml_threshold, "Charged", "Uncharged")
  )

ggplot(tbl_density, aes(x = charging_likelihood, fill = charging_status)) +
  geom_histogram(binwidth = 5, alpha = 0.8, position = "identity") +
  geom_vline(xintercept = ml_threshold, linetype = "dashed", linewidth = 1, color = "grey30") +
  scale_fill_manual(values = c("Charged" = "#E69F00", "Uncharged" = "#56B4E9")) +
  scale_y_continuous(labels = comma) +
  labs(x = "Charging Likelihood (ML tag, 0-255)", y = "Number of Reads", fill = "Classification") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid() +
  theme(legend.position = "top")
```

### Charging Summary

```{r}
#| label: charging-summary-table

charging_summary_tbl <- map_dfr(sample_list, function(s) {
  sample_prob <- charging_prob_all |> filter(sample == s)
  n_charged <- sum(sample_prob$charging_likelihood >= ml_threshold)
  n_uncharged <- sum(sample_prob$charging_likelihood < ml_threshold)
  n_total <- nrow(sample_prob)

  tibble(
    Sample = s,
    Charged = n_charged,
    Uncharged = n_uncharged,
    Total = n_total,
    `% Charged` = n_charged / n_total * 100,
    `% Uncharged` = n_uncharged / n_total * 100
  )
})

gt(charging_summary_tbl) |>
  fmt_number(columns = c(Charged, Uncharged, Total), use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = c(`% Charged`, `% Uncharged`), decimals = 1)
```

### Per-tRNA Abundance

```{r}
#| label: fig-trna-abundance
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Total CPM per tRNA isodecoder. Non-standard tRNAs shown in red."

tbl_abundance <- charging_cpm_all |>
  mutate(
    is_standard = is_standard_trna(tRNA),
    isodecoder = extract_isodecoder(tRNA)
  ) |>
  group_by(sample, isodecoder) |>
  summarise(
    total_counts = sum(counts_charged) + sum(counts_uncharged),
    total_cpm = sum(cpm_charged) + sum(cpm_uncharged),
    is_standard = first(is_standard),
    .groups = "drop"
  ) |>
  filter(total_counts >= 5) |>
  mutate(isodecoder = fct_reorder(isodecoder, total_cpm, .fun = mean))

ggplot(tbl_abundance, aes(x = isodecoder, y = total_cpm, fill = is_standard)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E41A1C"), guide = "none") +
  scale_y_continuous(labels = comma) +
  labs(x = NULL, y = "CPM") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
```

### Per-tRNA Charging

```{r}
#| label: fig-trna-charging
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Percent charged per tRNA isodecoder. Non-standard tRNAs shown in red."

tbl_charging <- charging_cpm_all |>
  mutate(
    is_standard = is_standard_trna(tRNA),
    isodecoder = extract_isodecoder(tRNA)
  ) |>
  group_by(sample, isodecoder) |>
  summarise(
    counts_charged = sum(counts_charged),
    counts_uncharged = sum(counts_uncharged),
    is_standard = first(is_standard),
    .groups = "drop"
  ) |>
  mutate(
    total_counts = counts_charged + counts_uncharged,
    pct_charged = counts_charged / total_counts * 100
  ) |>
  filter(total_counts >= 5) |>
  mutate(isodecoder = fct_reorder(isodecoder, pct_charged, .fun = mean))

ggplot(tbl_charging, aes(x = isodecoder, y = pct_charged, fill = is_standard)) +
  geom_col() +
  scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E41A1C"), guide = "none") +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
  labs(x = NULL, y = "% Charged") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
```

### Top tRNAs by Read Count

::: {.panel-tabset}

```{r}
#| label: top-trnas-tabs
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_top_trnas <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(isodecoder = extract_isodecoder(tRNA)) |>
    group_by(isodecoder) |>
    summarise(
      counts_charged = sum(counts_charged),
      counts_uncharged = sum(counts_uncharged),
      cpm_charged = sum(cpm_charged),
      cpm_uncharged = sum(cpm_uncharged),
      .groups = "drop"
    ) |>
    mutate(
      total_counts = counts_charged + counts_uncharged,
      total_cpm = cpm_charged + cpm_uncharged,
      pct_charged = counts_charged / total_counts * 100
    ) |>
    arrange(desc(total_counts)) |>
    head(20) |>
    dplyr::select(
      Isodecoder = isodecoder,
      `Charged` = counts_charged,
      `Uncharged` = counts_uncharged,
      `Total` = total_counts,
      `% Charged` = pct_charged,
      `CPM` = total_cpm
    )

  tbl <- gt(tbl_top_trnas) |>
    tab_header(title = paste("Top 20 tRNAs -", s)) |>
    fmt_number(columns = c(`Charged`, `Uncharged`, `Total`), use_seps = TRUE, decimals = 0) |>
    fmt_number(columns = `% Charged`, decimals = 1) |>
    fmt_number(columns = CPM, decimals = 0)

  cat(as_raw_html(tbl))
  cat("\n\n")
}
```

:::

## Base Calling Quality

### Error Frequency by Position

```{r}
#| label: fig-error-frequency
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Mean error frequency across all tRNAs by reference position."

tbl_error <- bcerror_all |>
  group_by(sample, Position) |>
  summarise(
    mean_mismatch = mean(MismatchFreq, na.rm = TRUE),
    mean_insertion = mean(InsertionFreq, na.rm = TRUE),
    mean_deletion = mean(DeletionFreq, na.rm = TRUE),
    .groups = "drop"
  )

tbl_error_long <- tbl_error |>
  dplyr::select(sample, Position, Mismatch = mean_mismatch, Insertion = mean_insertion, Deletion = mean_deletion) |>
  pivot_longer(cols = c(Mismatch, Insertion, Deletion), names_to = "Error Type", values_to = "Frequency")

ggplot(tbl_error_long, aes(x = Position, y = Frequency, color = `Error Type`)) +
  geom_line(linewidth = 0.8) +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  labs(x = "Position in Reference", y = "Error Frequency") +
  facet_wrap(~sample) +
  theme_minimal_hgrid() +
  theme(legend.position = "top")
```

### Coverage by Position

```{r}
#| label: fig-coverage-position
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Mean read coverage across all tRNAs by reference position."

tbl_coverage <- bcerror_all |>
  group_by(sample, Position) |>
  summarise(
    mean_coverage = mean(Spanning_Reads, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(tbl_coverage, aes(x = Position, y = mean_coverage)) +
  geom_area(fill = "#56B4E9", alpha = 0.6) +
  geom_line(color = "#0072B2", linewidth = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(x = "Position in Reference", y = "Mean Coverage (Reads)") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid()
```

## Summary

```{r}
#| label: summary-table
#| tbl-cap: "Charging classification summary across all samples."

summary_tbl <- map_dfr(sample_list, function(s) {
  sample_prob <- charging_prob_all |> filter(sample == s)
  n_charged <- sum(sample_prob$charging_likelihood >= ml_threshold)
  n_uncharged <- sum(sample_prob$charging_likelihood < ml_threshold)
  n_total <- nrow(sample_prob)

  tibble(
    Sample = s,
    `Total Classified Reads` = n_total,
    `Charged Reads` = n_charged,
    `Uncharged Reads` = n_uncharged,
    `% Charged` = n_charged / n_total * 100
  )
})

gt(summary_tbl) |>
  fmt_number(columns = c(`Total Classified Reads`, `Charged Reads`, `Uncharged Reads`), use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = `% Charged`, decimals = 1)
```

**ML threshold used:** `r ml_threshold`

{{< include _custom.qmd >}}

## Appendix {.appendix}

### Reference Sequence Similarity

This section shows pairwise sequence similarity across all sequences in the reference FASTA file. High similarity between sequences may indicate potential cross-mapping issues during alignment.

```{r}
#| label: reference-similarity-check
#| include: false

# Check if similarity matrix is available
has_similarity_matrix <- !is.null(similarity_matrix)
```

```{r}
#| label: fig-similarity-heatmap
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Pairwise sequence identity heatmap. Amino acid annotations shown on axes. High similarity (>90%) shown in orange/red."
#| eval: !expr has_similarity_matrix

# Parse tRNA names to extract amino acid and anticodon
parse_trna_name <- function(name) {
  parts <- str_split(name, "-")[[1]]
  if (length(parts) >= 3 && parts[1] == "tRNA") {
    aa <- parts[2]
    anticodon <- parts[3]
  } else if (length(parts) >= 2) {
    aa <- parts[1]
    anticodon <- if (length(parts) >= 2) parts[2] else "unknown"
  } else {
    aa <- "unknown"
    anticodon <- "unknown"
  }
  list(aa = aa, anticodon = anticodon)
}

# Parse all sequence names
parsed <- lapply(rownames(similarity_matrix), parse_trna_name)
aa_labels <- sapply(parsed, function(x) x$aa)

# Create color palette for amino acids
unique_aa <- unique(aa_labels)
n_aa <- length(unique_aa)

if (n_aa <= 20) {
  aa_colors <- setNames(rainbow(n_aa, s = 0.7, v = 0.8), unique_aa)
} else {
  aa_colors <- setNames(
    colorRampPalette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
                       "#FF7F00", "#FFFF33", "#A65628", "#F781BF"))(n_aa),
    unique_aa
  )
}

# Color scale for similarity values
col_fun <- colorRamp2(c(50, 75, 90, 100), c("white", "lightyellow", "orange", "red"))

# Create annotations
row_anno <- rowAnnotation(
  AA = aa_labels,
  col = list(AA = aa_colors),
  show_annotation_name = TRUE,
  annotation_name_side = "top"
)

col_anno <- HeatmapAnnotation(
  AA = aa_labels,
  col = list(AA = aa_colors),
  show_annotation_name = FALSE
)

# Draw heatmap
ht <- Heatmap(
  similarity_matrix,
  name = "% Identity",
  col = col_fun,
  left_annotation = row_anno,
  top_annotation = col_anno,
  show_row_names = nrow(similarity_matrix) <= 50,
  show_column_names = nrow(similarity_matrix) <= 50,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 6),
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",
  row_title = "Reference Sequences",
  column_title = "Reference Sequence Similarity"
)

draw(ht)
```

```{r}
#| label: no-similarity-matrix
#| eval: !expr (!has_similarity_matrix)
#| results: asis

cat("*Reference similarity matrix not available.*\n")
```

#### Highly Similar Sequence Pairs

Sequence pairs with >90% identity may be at risk for cross-mapping:

```{r}
#| label: high-similarity-pairs
#| eval: !expr has_similarity_matrix

high_sim_threshold <- 90

# Find pairs with high similarity (excluding diagonal)
pairs <- data.frame()
for (i in 1:(nrow(similarity_matrix) - 1)) {
  for (j in (i + 1):ncol(similarity_matrix)) {
    if (similarity_matrix[i, j] >= high_sim_threshold) {
      pairs <- rbind(pairs, data.frame(
        Seq1 = rownames(similarity_matrix)[i],
        Seq2 = colnames(similarity_matrix)[j],
        Percent_Identity = round(similarity_matrix[i, j], 2)
      ))
    }
  }
}

# Sort by similarity (descending)
pairs <- pairs |> arrange(desc(Percent_Identity))

if (nrow(pairs) > 0) {
  gt(head(pairs, 50)) |>
    cols_label(
      Seq1 = "Sequence 1",
      Seq2 = "Sequence 2",
      Percent_Identity = "% Identity"
    ) |>
    fmt_number(columns = Percent_Identity, decimals = 1)
} else {
  cat(paste0("No sequence pairs found with >= ", high_sim_threshold, "% identity.\n"))
}
```

#### Similarity Distribution

```{r}
#| label: fig-similarity-histogram
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Distribution of pairwise sequence similarities. Red dashed line indicates 90% threshold."
#| eval: !expr has_similarity_matrix

# Get upper triangle values (excluding diagonal)
upper_tri <- similarity_matrix[upper.tri(similarity_matrix)]

tibble(similarity = upper_tri) |>
  ggplot(aes(x = similarity)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 90, linetype = "dashed", color = "red", linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(x = "Percent Identity", y = "Number of Pairs") +
  theme_minimal_hgrid()
```

```{r}
#| label: similarity-summary-stats
#| eval: !expr has_similarity_matrix

upper_tri <- similarity_matrix[upper.tri(similarity_matrix)]

tibble(
  Statistic = c("Minimum", "Median", "Mean", "Maximum", "Pairs >= 90%", "Pairs >= 95%", "Total pairs"),
  Value = c(
    sprintf("%.1f%%", min(upper_tri)),
    sprintf("%.1f%%", median(upper_tri)),
    sprintf("%.1f%%", mean(upper_tri)),
    sprintf("%.1f%%", max(upper_tri)),
    as.character(sum(upper_tri >= 90)),
    as.character(sum(upper_tri >= 95)),
    as.character(length(upper_tri))
  )
) |>
  gt()
```

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info

sessioninfo::session_info()
```

:::
