---
title: "aa-tRNA-seq QC Report"
author: "Pipeline"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
params:
  sample: "sample1"
  output_dir: ".tests/outputs"
  ml_threshold: 200
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(paletteer)
library(fs)
library(patchwork)
library(scales)

# Parameters
sample <- params$sample
output_dir <- params$output_dir
ml_threshold <- params$ml_threshold

# Paths to input files
tables_dir <- fs::path(output_dir, "summary", "tables", sample)

align_stats_path <- fs::path(tables_dir, paste0(sample, ".align_stats.tsv.gz"))
charging_prob_path <- fs::path(tables_dir, paste0(sample, ".charging_prob.tsv.gz"))
charging_cpm_path <- fs::path(tables_dir, paste0(sample, ".charging.cpm.tsv.gz"))
bcerror_path <- fs::path(tables_dir, paste0(sample, ".bcerror.tsv.gz"))
```

## Sample: `r sample`

This report summarizes QC metrics for sample **`r sample`** from the aa-tRNA-seq pipeline.

## Alignment Summary

```{r}
#| label: load-align-stats

align_stats <- read_tsv(align_stats_path, show_col_types = FALSE)
```

### Read Progression Through Pipeline

```{r}
#| label: tbl-alignment-summary
#| tbl-cap: "Read counts at each processing stage"

tbl_align <- align_stats |>
  dplyr::select(
    Stage = info,
    `Total Reads` = n_reads,
    `Mapped Reads` = mapped_reads,
    `% Mapped` = pct_mapped,
    `Mean Length` = mean_length,
    `Median Length` = median_length,
    `Mean Quality` = mean_base_quality,
    `Mean MAPQ` = mean_MAPQ
  ) |>
  mutate(
    Stage = factor(Stage, levels = c("unmapped", "aligned", "classified")),
    Stage = recode(Stage,
      "unmapped" = "Rebasecalled",
      "aligned" = "Aligned",
      "classified" = "Classified"
    )
  )

gt(tbl_align) |>
  fmt_number(columns = c(`Mean Length`, `Mean Quality`, `Mean MAPQ`), decimals = 1) |>
  fmt_number(columns = `% Mapped`, decimals = 1) |>
  opt_interactive(use_search = TRUE, use_highlight = TRUE, use_sorting = TRUE)
```

### Alignment Funnel

```{r}
#| label: fig-alignment-funnel
#| fig-cap: "Read counts at each processing stage showing retention through the pipeline"
#| fig-width: 8
#| fig-height: 5

tbl_funnel <- align_stats |>
  dplyr::select(info, n_reads) |>
  mutate(
    stage = factor(info, levels = c("unmapped", "aligned", "classified")),
    stage = recode(stage,
      "unmapped" = "Rebasecalled",
      "aligned" = "Aligned",
      "classified" = "Classified"
    )
  )

ggplot(tbl_funnel, aes(x = stage, y = n_reads, fill = stage)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = comma(n_reads)), vjust = -0.5, size = 4) +
  scale_fill_paletteer_d("ggokabeito::okabe_ito") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(
    x = "Processing Stage",
    y = "Number of Reads"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## Charging Classification

```{r}
#| label: load-charging-data

charging_prob <- read_tsv(charging_prob_path, show_col_types = FALSE)
charging_cpm <- read_tsv(charging_cpm_path, show_col_types = FALSE)
```

### Charging Likelihood Distribution

The Remora model assigns a charging likelihood score (0-255) to each read. Reads with scores >= `r ml_threshold` are classified as charged.

```{r}
#| label: fig-charging-density
#| fig-cap: !expr paste0("Distribution of charging likelihood scores. Threshold = ", ml_threshold, " (dashed line)")
#| fig-width: 8
#| fig-height: 5

tbl_density <- charging_prob |>
  mutate(
    charging_status = if_else(charging_likelihood >= ml_threshold, "Charged", "Uncharged")
  )

ggplot(tbl_density, aes(x = charging_likelihood, fill = charging_status)) +
  geom_histogram(binwidth = 5, alpha = 0.8, position = "identity") +
  geom_vline(xintercept = ml_threshold, linetype = "dashed", linewidth = 1, color = "grey30") +
  scale_fill_manual(values = c("Charged" = "#E69F00", "Uncharged" = "#56B4E9")) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Charging Likelihood (ML tag, 0-255)",
    y = "Number of Reads",
    fill = "Classification"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Charging Summary

```{r}
#| label: tbl-charging-summary
#| tbl-cap: "Summary of charged vs uncharged reads"

n_charged <- sum(charging_prob$charging_likelihood >= ml_threshold)
n_uncharged <- sum(charging_prob$charging_likelihood < ml_threshold)
n_total <- nrow(charging_prob)

tbl_charging_summary <- tibble(
  Status = c("Charged", "Uncharged", "Total"),
  `Read Count` = c(n_charged, n_uncharged, n_total),
  Percentage = c(
    n_charged / n_total * 100,
    n_uncharged / n_total * 100,
    100
  )
)

gt(tbl_charging_summary) |>
  fmt_number(columns = Percentage, decimals = 1) |>
  fmt_number(columns = `Read Count`, use_seps = TRUE, decimals = 0)
```

### Per-tRNA Charging

```{r}
#| label: fig-trna-charging-heatmap
#| fig-cap: "Charging status by tRNA isodecoder (CPM normalized)"
#| fig-width: 8
#| fig-height: 6

# Filter to tRNAs with at least some reads
tbl_heatmap <- charging_cpm |>
  mutate(
    total_counts = counts_charged + counts_uncharged,
    pct_charged = counts_charged / total_counts * 100
  ) |>
  dplyr::filter(total_counts >= 5) |>
  arrange(desc(pct_charged))

if (nrow(tbl_heatmap) > 0) {
  tbl_heatmap <- tbl_heatmap |>
    mutate(tRNA = fct_reorder(tRNA, pct_charged))

  ggplot(tbl_heatmap, aes(x = 1, y = tRNA, fill = pct_charged)) +
    geom_tile() +
    geom_text(aes(label = paste0(round(pct_charged, 0), "%")), size = 3) +
    scale_fill_distiller(palette = "RdYlBu", direction = -1, limits = c(0, 100)) +
    labs(
      x = NULL,
      y = "tRNA Isodecoder",
      fill = "% Charged"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      panel.grid = element_blank()
    )
} else {
  cat("No tRNAs with >= 5 reads")
}
```

### Top tRNAs by Read Count

```{r}
#| label: tbl-top-trnas
#| tbl-cap: "Top 20 tRNAs by total read count"

tbl_top_trnas <- charging_cpm |>
  mutate(
    total_counts = counts_charged + counts_uncharged,
    total_cpm = cpm_charged + cpm_uncharged,
    pct_charged = counts_charged / total_counts * 100
  ) |>
  arrange(desc(total_counts)) |>
  head(20) |>
  dplyr::select(
    tRNA,
    `Charged` = counts_charged,
    `Uncharged` = counts_uncharged,
    `Total` = total_counts,
    `% Charged` = pct_charged,
    `CPM` = total_cpm
  )

gt(tbl_top_trnas) |>
  fmt_number(columns = c(`Charged`, `Uncharged`, `Total`), use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = `% Charged`, decimals = 1) |>
  fmt_number(columns = CPM, decimals = 0) |>
  opt_interactive(use_search = TRUE, use_highlight = TRUE, use_sorting = TRUE)
```

## Base Calling Quality

```{r}
#| label: load-bcerror

bcerror <- read_tsv(bcerror_path, show_col_types = FALSE)
```

### Error Frequency by Position

```{r}
#| label: fig-error-frequency
#| fig-cap: "Base calling error frequency across reference positions"
#| fig-width: 10
#| fig-height: 5

# Aggregate across all references
tbl_error <- bcerror |>
  group_by(Position) |>
  summarise(
    mean_bcerror = mean(BCErrorFreq, na.rm = TRUE),
    mean_mismatch = mean(MismatchFreq, na.rm = TRUE),
    mean_insertion = mean(InsertionFreq, na.rm = TRUE),
    mean_deletion = mean(DeletionFreq, na.rm = TRUE),
    mean_coverage = mean(Spanning_Reads, na.rm = TRUE),
    .groups = "drop"
  )

tbl_error_long <- tbl_error |>
  dplyr::select(Position, Mismatch = mean_mismatch, Insertion = mean_insertion, Deletion = mean_deletion) |>
  pivot_longer(cols = -Position, names_to = "Error Type", values_to = "Frequency")

ggplot(tbl_error_long, aes(x = Position, y = Frequency, color = `Error Type`)) +
  geom_line(linewidth = 0.8) +
  scale_color_paletteer_d("ggokabeito::okabe_ito") +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  labs(
    x = "Position in Reference",
    y = "Error Frequency"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```

### Coverage by Position

```{r}
#| label: fig-coverage-position
#| fig-cap: "Mean read coverage across reference positions"
#| fig-width: 10
#| fig-height: 4

ggplot(tbl_error, aes(x = Position, y = mean_coverage)) +
  geom_area(fill = "#56B4E9", alpha = 0.6) +
  geom_line(color = "#0072B2", linewidth = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(
    x = "Position in Reference",
    y = "Mean Coverage (Reads)"
  ) +
  theme_minimal()
```

::: {.callout-note}
## Summary

- **Total classified reads**: `r comma(nrow(charging_prob))`
- **Charged reads**: `r comma(n_charged)` (`r round(n_charged/n_total*100, 1)`%)
- **Uncharged reads**: `r comma(n_uncharged)` (`r round(n_uncharged/n_total*100, 1)`%)
- **ML threshold**: `r ml_threshold`

:::

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info

sessioninfo::session_info()
```

:::
