---
title: "aa-tRNA-seq Combined QC Report"
author: "Pipeline"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
params:
  config_file: "config/config.yml"
  ml_threshold: 200
---

## Pipeline Overview

```{mermaid}
flowchart LR
  A[POD5 files] --> B[Merge POD5s]
  B --> C[Rebasecall<br>Dorado]
  C --> D[FASTQ]
  D --> E[Align<br>BWA MEM]
  E --> F[Classify Charging<br>escpod]
  F --> G[Transfer Tags]
  G --> H[Add Adapter Tags]
  H --> I[Summary Tables<br>& QC Metrics]
```

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(paletteer)
library(fs)
library(patchwork)
library(scales)
library(yaml)
library(cowplot)
library(ggrepel)
library(ComplexHeatmap)
library(circlize)

# Read pipeline config
config <- read_yaml(params$config_file)
output_dir <- fs::path_abs(config$output_directory, start = dirname(params$config_file))
ml_threshold <- params$ml_threshold

# escapepod-models names the LDX barcodes `nbc01`..`nbc16`, and that is what the
# model emits into classifications.csv and demux_summary.tsv.gz -- but LDX is the
# name this project uses for them. Normalise both sides of every barcode
# join/display so the report speaks our vocabulary and still matches the files.
# Pure documented rename; WDX barcodes pass through untouched. Defined at top
# level because the demux chunk below uses it regardless of the samples format.
normalize_barcode <- function(x) {
  ifelse(!is.na(x) & grepl("^nbc[0-9]+$", x), sub("^nbc", "ldx", x), x)
}

# Parse samples from config
# Supports both YAML format (with runs) and TSV format
if (!is.null(config$samples))
{
  samples_path <- fs::path_abs(config$samples, start = dirname(params$config_file))
  if (fs::path_ext(samples_path) %in% c("yml", "yaml")) {
    samples_config <- read_yaml(samples_path)
    sample_metadata <- samples_config$runs |>
      purrr::map_dfr(function(run) {
        purrr::imap_dfr(run$samples, function(val, name) {
          if (is.list(val)) {
            # Dict format: wdx/ldx + edx. `wdx:` and `ldx:` are alternative
            # spellings of the same field -- the two demux backends are mutually
            # exclusive -- so coalesce them rather than reading only wdx.
            tibble(
              sample = name,
              barcode = val$wdx %||% val$ldx %||% NA_character_,
              edx = val$edx %||% NA_character_,
              run_path = run$path,
              barcode_kit = run$barcode_kit %||% NA_character_
            )
          } else {
            # Simple format: barcode string or null
            tibble(
              sample = name,
              barcode = val %||% NA_character_,
              edx = NA_character_,
              run_path = run$path,
              barcode_kit = run$barcode_kit %||% NA_character_
            )
          }
        })
      })
    sample_metadata <- sample_metadata |>
      mutate(barcode = normalize_barcode(barcode))
    sample_list <- sample_metadata$sample
  } else {
    # TSV format: first column is sample name, second is run directory
    sample_metadata <- read_tsv(samples_path, col_names = c("sample", "run_path"),
                                show_col_types = FALSE) |>
      mutate(run_path = fs::path_abs(run_path, start = dirname(samples_path))) |>
      group_by(sample) |>
      summarise(run_path = paste(run_path, collapse = "; "), .groups = "drop")
    sample_list <- sample_metadata$sample
  }
} else {
 stop("No samples defined in config file")
}

# Apply sample ordering from config if specified
if (!is.null(config$report$sample_order)) {
  specified_order <- config$report$sample_order
  # Put specified samples first (in order), then any remaining samples
  sample_list <- c(
    intersect(specified_order, sample_list),
    setdiff(sample_list, specified_order)
  )
}

# Extract isodecoder from gtrnadb-format tRNA names
# Pattern: Genus_species_tRNA-Aaa-NNN-X-Y → Aaa-NNN (handles iMet with 4 chars)
# Returns list with isodecoder name and whether it matched the tRNA pattern
extract_isodecoder <- function(trna_name) {
  isodecoder <- str_extract(trna_name, "(?<=tRNA-)[A-Za-z]{3,4}[0-9]?-[ACGT]{3}")
  if_else(is.na(isodecoder), trna_name, isodecoder)
}

# Check if name matches standard tRNA isodecoder pattern
is_standard_trna <- function(trna_name) {
 str_detect(trna_name, "tRNA-[A-Za-z]{3,4}[0-9]?-[ACGT]{3}")
}
```

This report summarizes QC metrics for **`r length(sample_list)` samples** (`r paste(sample_list, collapse = ", ")`) from the aa-tRNA-seq pipeline. See [Sample Details](#sample-details) for run paths and barcodes.

```{r}
#| label: load-all-data
#| include: false

# Load all data upfront
align_stats_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".align_stats.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

charging_prob_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".charging_prob.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
}) |>
  filter(!str_detect(tRNA, "tRNA-Und"))

charging_cpm_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".charging.cpm.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
}) |>
  filter(!str_detect(tRNA, "tRNA-Und"))

bcerror_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".bcerror.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

odds_ratios_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  or_path <- fs::path(tables_dir, paste0(s, ".odds_ratios.tsv.gz"))
  if (fs::file_exists(or_path)) {
    read_tsv(or_path, show_col_types = FALSE) |> mutate(sample = s)
  } else {
    tibble()
  }
})

has_odds_ratios <- nrow(odds_ratios_all) > 0
if (has_odds_ratios) {
  odds_ratios_all <- odds_ratios_all |>
    mutate(sample = factor(sample, levels = sample_list))
}

# Convert sample to factor with correct ordering for all datasets
align_stats_all <- align_stats_all |> mutate(sample = factor(sample, levels = sample_list))
charging_prob_all <- charging_prob_all |> mutate(sample = factor(sample, levels = sample_list))
charging_cpm_all <- charging_cpm_all |> mutate(sample = factor(sample, levels = sample_list))
bcerror_all <- bcerror_all |> mutate(sample = factor(sample, levels = sample_list))

# Load reference similarity matrix
similarity_matrix_path <- fs::path(output_dir, "summary", "qc", "reference_similarity.tsv")
similarity_matrix <- NULL
if (fs::file_exists(similarity_matrix_path)) {
  similarity_df <- read_tsv(similarity_matrix_path, show_col_types = FALSE)
  similarity_matrix <- as.matrix(similarity_df[, -1])
  rownames(similarity_matrix) <- similarity_df[[1]]
}
```

```{r}
#| label: plot-helpers
#| include: false

plot_alignment_funnel <- function(s) {
  tbl_funnel <- align_stats_all |>
    filter(sample == s) |>
    dplyr::select(sample, info, n_reads) |>
    mutate(
      stage = factor(info, levels = c("unmapped", "aligned", "classified")),
      stage = recode(stage,
        "unmapped" = "Rebasecalled",
        "aligned" = "Aligned",
        "classified" = "Classified"
      )
    )

  ggplot(tbl_funnel, aes(x = stage, y = n_reads, fill = stage)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = label_number(scale_cut = cut_short_scale(), accuracy = 0.1)(n_reads)),
              vjust = -0.5, size = 3) +
    scale_fill_brewer(palette = "Set2") +
    scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                       expand = expansion(mult = c(0, 0.15))) +
    labs(x = "Processing Stage", y = "Number of Reads") +
    ggtitle("Read Progression") +
    theme_minimal_hgrid() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
}

plot_charging_density <- function(s) {
  tbl_density <- charging_prob_all |>
    filter(sample == s) |>
    mutate(
      charging_status = if_else(charging_likelihood >= ml_threshold, "Charged", "Uncharged")
    )

  ggplot(tbl_density, aes(x = charging_likelihood, fill = charging_status)) +
    geom_histogram(binwidth = 5, alpha = 0.8, position = "identity") +
    geom_vline(xintercept = ml_threshold, linetype = "dashed", linewidth = 1, color = "grey30") +
    scale_fill_manual(values = c("Charged" = "#E69F00", "Uncharged" = "#56B4E9")) +
    scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
    labs(x = "Charging Likelihood (ML tag, 0-255)", y = "Number of Reads", fill = "Classification") +
    ggtitle("Charging Score Distribution") +
    theme_minimal_hgrid() +
    theme(legend.position = "top")
}

plot_trna_abundance <- function(s) {
  tbl_abundance <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(
      is_standard = is_standard_trna(tRNA),
      isodecoder = extract_isodecoder(tRNA)
    ) |>
    group_by(isodecoder) |>
    summarise(
      total_counts = sum(counts_charged) + sum(counts_uncharged),
      total_cpm = sum(cpm_charged) + sum(cpm_uncharged),
      is_standard = first(is_standard),
      .groups = "drop"
    ) |>
    filter(total_counts >= 5) |>
    mutate(isodecoder = fct_reorder(isodecoder, total_cpm))

  ggplot(tbl_abundance, aes(x = isodecoder, y = total_cpm, fill = is_standard)) +
    geom_col() +
    scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E41A1C"), guide = "none") +
    scale_y_continuous(labels = comma) +
    labs(x = NULL, y = "CPM") +
    ggtitle("tRNA Abundance (CPM)") +
    theme_minimal_hgrid() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
}

plot_trna_charging <- function(s) {
  tbl_charging <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(
      is_standard = is_standard_trna(tRNA),
      isodecoder = extract_isodecoder(tRNA)
    ) |>
    group_by(isodecoder) |>
    summarise(
      counts_charged = sum(counts_charged),
      counts_uncharged = sum(counts_uncharged),
      is_standard = first(is_standard),
      .groups = "drop"
    ) |>
    mutate(
      total_counts = counts_charged + counts_uncharged,
      pct_charged = counts_charged / total_counts * 100
    ) |>
    filter(total_counts >= 5) |>
    mutate(isodecoder = fct_reorder(isodecoder, pct_charged))

  ggplot(tbl_charging, aes(x = isodecoder, y = pct_charged, fill = is_standard)) +
    geom_col() +
    scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E41A1C"), guide = "none") +
    scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
    labs(x = NULL, y = "% Charged") +
    ggtitle("tRNA Charging (%)") +
    theme_minimal_hgrid() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)))
}

plot_error_frequency <- function(s) {
  tbl_error <- bcerror_all |>
    filter(sample == s) |>
    group_by(Position) |>
    summarise(
      mean_mismatch = mean(MismatchFreq, na.rm = TRUE),
      mean_insertion = mean(InsertionFreq, na.rm = TRUE),
      mean_deletion = mean(DeletionFreq, na.rm = TRUE),
      .groups = "drop"
    )

  tbl_error_long <- tbl_error |>
    dplyr::select(Position, Mismatch = mean_mismatch,
                  Insertion = mean_insertion, Deletion = mean_deletion) |>
    pivot_longer(cols = c(Mismatch, Insertion, Deletion),
                 names_to = "Error Type", values_to = "Frequency")

  ggplot(tbl_error_long, aes(x = Position, y = Frequency, color = `Error Type`)) +
    geom_line(linewidth = 0.8) +
    scale_color_brewer(palette = "Set2") +
    scale_y_continuous(labels = percent_format(scale = 100)) +
    labs(x = "Position in Reference", y = "Error Frequency") +
    ggtitle("Error Frequency") +
    theme_minimal_hgrid() +
    theme(legend.position = "top")
}

plot_coverage_position <- function(s) {
  tbl_coverage <- bcerror_all |>
    filter(sample == s) |>
    group_by(Position) |>
    summarise(
      mean_coverage = mean(Spanning_Reads, na.rm = TRUE),
      .groups = "drop"
    )

  ggplot(tbl_coverage, aes(x = Position, y = mean_coverage)) +
    geom_area(fill = "#56B4E9", alpha = 0.6) +
    geom_line(color = "#0072B2", linewidth = 0.5) +
    scale_y_continuous(labels = label_number(scale_cut = cut_short_scale())) +
    labs(x = "Position in Reference", y = "Mean Coverage (Reads)") +
    ggtitle("Position Coverage") +
    theme_minimal_hgrid()
}

plot_odds_ratio_volcano <- function(s, n_label = 15) {
  or_data <- odds_ratios_all |>
    filter(sample == s, is.finite(log_odds_ratio), !is.na(p_adjusted))

  if (nrow(or_data) == 0) {
    return(ggplot() + annotate("text", x = 0.5, y = 0.5, label = "No odds ratio data") + theme_void())
  }

  or_data <- or_data |>
    mutate(
      neg_log10_padj = -log10(p_adjusted),
      significant = p_adjusted < 0.05,
      involves_charging = str_detect(pos1, "charging") | str_detect(pos2, "charging"),
      pair_label = paste0(pos1, " : ", pos2)
    )

  # Select top points to label (by significance and effect size)
  top_pts <- or_data |>
    filter(significant) |>
    arrange(p_adjusted) |>
    head(n_label)

  p <- ggplot(or_data, aes(x = log_odds_ratio, y = neg_log10_padj)) +
    geom_point(aes(color = significant, shape = involves_charging), alpha = 0.6, size = 1.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
    scale_color_manual(values = c("TRUE" = "#D55E00", "FALSE" = "grey70"),
                       labels = c("TRUE" = "p_adj < 0.05", "FALSE" = "NS")) +
    scale_shape_manual(values = c("TRUE" = 17, "FALSE" = 16),
                       labels = c("TRUE" = "Yes", "FALSE" = "No")) +
    labs(x = "Log Odds Ratio", y = expression(-log[10](p[adjusted])),
         color = "Significant", shape = "Involves Charging") +
    ggtitle("Pairwise Modification Odds Ratios") +
    theme_minimal_hgrid() +
    theme(legend.position = "top")

  if (nrow(top_pts) > 0) {
    p <- p + ggrepel::geom_text_repel(
      data = top_pts,
      aes(label = pair_label),
      size = 2.5, max.overlaps = 20, segment.color = "grey50"
    )
  }

  p
}

plot_odds_ratio_heatmap <- function(s) {
  or_data <- odds_ratios_all |>
    filter(sample == s, p_adjusted < 0.05, is.finite(log_odds_ratio))

  if (nrow(or_data) < 3) {
    return(
      ggplot() +
        annotate("text", x = 0.5, y = 0.5,
                 label = "Fewer than 3 significant pairs;\nheatmap not shown.",
                 size = 4, color = "grey40") +
        theme_void()
    )
  }

  # Build symmetric matrix of log odds ratios
  positions <- sort(unique(c(or_data$pos1, or_data$pos2)))

  mat <- matrix(0, nrow = length(positions), ncol = length(positions),
                dimnames = list(positions, positions))

  for (i in seq_len(nrow(or_data))) {
    r <- or_data[i, ]
    mat[r$pos1, r$pos2] <- r$log_odds_ratio
    mat[r$pos2, r$pos1] <- r$log_odds_ratio
  }

  # Convert to long format for ggplot
  mat_df <- as.data.frame(mat) |>
    tibble::rownames_to_column("row") |>
    pivot_longer(-row, names_to = "col", values_to = "log_or") |>
    filter(row != col) |>
    mutate(
      row = factor(row, levels = positions),
      col = factor(col, levels = positions)
    )

  max_abs <- max(abs(mat_df$log_or), na.rm = TRUE)

  ggplot(mat_df, aes(x = col, y = row, fill = log_or)) +
    geom_tile(color = "white", linewidth = 0.3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, limits = c(-max_abs, max_abs),
                         name = "Log OR") +
    labs(x = NULL, y = NULL) +
    ggtitle("Significant Pairwise Log Odds Ratios") +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.7)),
      axis.text.y = element_text(size = rel(0.7))
    )
}

plot_demux_funnel <- function(s, funnel_data) {
  tbl <- funnel_data |> filter(sample == s)
  if (nrow(tbl) == 0) {
    return(ggplot() + annotate("text", x = 0.5, y = 0.5, label = "No data") + theme_void())
  }

  # Compute % of previous stage
  tbl <- tbl |>
    arrange(stage_order) |>
    mutate(
      pct_of_prev = if_else(
        row_number() == 1, NA_real_,
        reads / lag(reads) * 100
      ),
      label = if_else(
        is.na(pct_of_prev),
        label_number(scale_cut = cut_short_scale(), accuracy = 0.1)(reads),
        paste0(
          label_number(scale_cut = cut_short_scale(), accuracy = 0.1)(reads),
          "\n(", round(pct_of_prev, 1), "% of prev)"
        )
      )
    )

  ggplot(tbl, aes(x = stage, y = reads, fill = stage)) +
    geom_col(width = 0.7) +
    geom_text(aes(label = label), vjust = -0.3, size = 3, lineheight = 0.85) +
    scale_fill_brewer(palette = "Blues") +
    scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                       expand = expansion(mult = c(0, 0.2))) +
    labs(x = NULL, y = "Number of Reads") +
    ggtitle(paste("Demux Yield Funnel —", s)) +
    theme_minimal_hgrid() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
}

make_sample_qc_panel <- function(s) {
  p_a <- plot_alignment_funnel(s)
  p_b <- plot_charging_density(s)
  p_c <- plot_trna_abundance(s)
  p_d <- plot_trna_charging(s)
  p_e <- plot_error_frequency(s)
  p_f <- plot_coverage_position(s)

  design <- "
AABB
CCCC
DDDD
EEFF
"

  p_a + p_b + p_c + p_d + p_e + p_f +
    plot_layout(design = design, heights = c(1, 1.3, 1.3, 1)) +
    plot_annotation(tag_levels = "A")
}
```

## Alignment Summary

### Read Progression Through Pipeline

::: {.panel-tabset}

```{r}
#| label: alignment-summary-tabs
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_align <- align_stats_all |>
    filter(sample == s) |>
    dplyr::select(
      Stage = info,
      `Total Reads` = n_reads,
      `Mapped Reads` = mapped_reads,
      `% Mapped` = pct_mapped,
      `Mean Length` = mean_length,
      `Median Length` = median_length,
      `Mean Quality` = mean_base_quality,
      `Mean MAPQ` = mean_MAPQ
    ) |>
    mutate(
      Stage = factor(Stage, levels = c("unmapped", "aligned", "classified")),
      Stage = recode(Stage,
        "unmapped" = "Rebasecalled",
        "aligned" = "Aligned",
        "classified" = "Classified"
      )
    )

  tbl <- gt(tbl_align) |>
    fmt_number(columns = c(`Total Reads`, `Mapped Reads`), suffixing = TRUE, decimals = 1) |>
    fmt_number(columns = c(`Mean Length`, `Mean Quality`, `Mean MAPQ`), decimals = 1) |>
    fmt_number(columns = `% Mapped`, decimals = 1)

  cat(as_raw_html(tbl))
  cat("\n\n")
}
```

:::

## Per-Sample QC Overview

Each tab shows a combined QC panel for one sample: (A) read progression through the pipeline, (B) charging score distribution, (C) tRNA abundance, (D) tRNA charging fraction, (E) base calling error frequency, and (F) position coverage.

::: {.panel-tabset}

```{r}
#| label: sample-qc-panels
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")
  res <- knitr::knit_child(text = c(
    "```{r}",
    "#| echo: false",
    "#| fig-width: 14",
    "#| fig-height: 18",
    sprintf("make_sample_qc_panel('%s')", s),
    "```",
    ""
  ), envir = environment(), quiet = TRUE)
  cat(res, sep = "\n")
  cat("\n\n")
}
```

:::

## Charging Classification

### Charging Summary

```{r}
#| label: charging-summary-table

charging_summary_tbl <- map_dfr(sample_list, function(s) {
  sample_prob <- charging_prob_all |> filter(sample == s)
  n_charged <- sum(sample_prob$charging_likelihood >= ml_threshold)
  n_uncharged <- sum(sample_prob$charging_likelihood < ml_threshold)
  n_total <- nrow(sample_prob)

  tibble(
    Sample = s,
    Charged = n_charged,
    Uncharged = n_uncharged,
    Total = n_total,
    `% Charged` = n_charged / n_total * 100,
    `% Uncharged` = n_uncharged / n_total * 100
  )
})

gt(charging_summary_tbl) |>
  fmt_number(columns = c(Charged, Uncharged, Total), suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = c(`% Charged`, `% Uncharged`), decimals = 1)
```

### Top tRNAs by Read Count

::: {.panel-tabset}

```{r}
#| label: top-trnas-tabs
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_top_trnas <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(isodecoder = extract_isodecoder(tRNA)) |>
    group_by(isodecoder) |>
    summarise(
      counts_charged = sum(counts_charged),
      counts_uncharged = sum(counts_uncharged),
      cpm_charged = sum(cpm_charged),
      cpm_uncharged = sum(cpm_uncharged),
      .groups = "drop"
    ) |>
    mutate(
      total_counts = counts_charged + counts_uncharged,
      total_cpm = cpm_charged + cpm_uncharged,
      pct_charged = counts_charged / total_counts * 100
    ) |>
    arrange(desc(total_counts)) |>
    head(20) |>
    dplyr::select(
      Isodecoder = isodecoder,
      `Charged` = counts_charged,
      `Uncharged` = counts_uncharged,
      `Total` = total_counts,
      `% Charged` = pct_charged,
      `CPM` = total_cpm
    )

  tbl <- gt(tbl_top_trnas) |>
    tab_header(title = paste("Top 20 tRNAs -", s)) |>
    fmt_number(columns = c(`Charged`, `Uncharged`, `Total`), suffixing = TRUE, decimals = 1) |>
    fmt_number(columns = `% Charged`, decimals = 1) |>
    fmt_number(columns = CPM, decimals = 0)

  cat(as_raw_html(tbl))
  cat("\n\n")
}
```

:::

## Modification Odds Ratios {.unnumbered}

```{r}
#| label: odds-ratios-section-guard
#| include: false
#| eval: !expr has_odds_ratios
```

This section shows pairwise odds ratios between modification positions (and charging status) for each sample. Odds ratios quantify co-occurrence patterns: a log OR > 0 indicates that modifications at two positions tend to co-occur on the same reads, while log OR < 0 indicates they tend to be mutually exclusive. P-values are FDR-adjusted (Benjamini-Hochberg).

::: {.panel-tabset}

```{r}
#| label: odds-ratios-tabs
#| results: asis
#| eval: !expr has_odds_ratios

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  # Volcano plot
  res_volcano <- knitr::knit_child(text = c(
    "```{r}",
    "#| echo: false",
    "#| fig-width: 10",
    "#| fig-height: 7",
    sprintf("plot_odds_ratio_volcano('%s')", s),
    "```",
    ""
  ), envir = environment(), quiet = TRUE)
  cat(res_volcano, sep = "\n")
  cat("\n\n")

  # Top 20 significant pairs table
  or_top <- odds_ratios_all |>
    dplyr::filter(sample == s, !is.na(p_adjusted)) |>
    dplyr::arrange(p_adjusted) |>
    dplyr::head(20) |>
    dplyr::select(
      `Position 1` = pos1,
      `Position 2` = pos2,
      `Log OR` = log_odds_ratio,
      `SE` = se_log_or,
      `Fisher OR` = fisher_or,
      `p (adjusted)` = p_adjusted,
      `Total Obs` = total_obs
    )

  if (nrow(or_top) > 0) {
    tbl <- gt(or_top) |>
      tab_header(title = paste("Top 20 Pairs by Significance -", s)) |>
      fmt_number(columns = c(`Log OR`, `SE`, `Fisher OR`), decimals = 2) |>
      fmt_scientific(columns = `p (adjusted)`, decimals = 2) |>
      fmt_number(columns = `Total Obs`, decimals = 0)
    cat(as_raw_html(tbl))
  } else {
    cat("*No significant pairs found.*\n")
  }
  cat("\n\n")

  # Heatmap
  res_heatmap <- knitr::knit_child(text = c(
    "```{r}",
    "#| echo: false",
    "#| fig-width: 10",
    "#| fig-height: 8",
    sprintf("plot_odds_ratio_heatmap('%s')", s),
    "```",
    ""
  ), envir = environment(), quiet = TRUE)
  cat(res_heatmap, sep = "\n")
  cat("\n\n")
}
```

:::

## Demultiplexing {.unnumbered}

```{r}
#| label: demux-data-load
#| include: false

# Check for demux summary files. Both backends write the same
# demux_summary.tsv.gz schema (predicted_barcode / n_reads / pct), so
# everything below is backend-neutral once the file is found.
demux_backend <- if (isTRUE(config$ldx$enabled)) {
  "LDX"
} else if (isTRUE(config$warpdemux$enabled)) {
  "WDX"
} else {
  NA_character_
}
has_demux <- !is.na(demux_backend)

demux_summary_all <- tibble()
if (has_demux) {
  # Find run IDs from sample metadata. Keyed on having a barcode, NOT on
  # barcode_kit: LDX deliberately configures no kit (the model bundle is
  # self-describing), so filtering on it finds nothing on an LDX run.
  run_ids <- unique(
    sample_metadata |>
      filter(!is.na(barcode)) |>
      pull(run_path) |>
      basename()
  )

  demux_summary_all <- map_dfr(run_ids, function(rid) {
    summary_path <- fs::path(output_dir, "demux", "read_ids", rid, "demux_summary.tsv.gz")
    if (fs::file_exists(summary_path)) {
      read_tsv(summary_path, show_col_types = FALSE) |>
        mutate(run_id = rid, predicted_barcode = normalize_barcode(predicted_barcode))
    } else {
      tibble()
    }
  })
}
has_demux_summary <- nrow(demux_summary_all) > 0

# Check for EDX concordance
has_edx <- "edx" %in% colnames(sample_metadata) && any(!is.na(sample_metadata$edx))
edx_concordance <- tibble()
if (has_edx) {
  edx_path <- fs::path(output_dir, "summary", "edx", "edx_concordance.tsv.gz")
  if (fs::file_exists(edx_path)) {
    edx_concordance <- read_tsv(edx_path, show_col_types = FALSE)
  }
}
has_edx_concordance <- nrow(edx_concordance) > 0

has_any_demux <- has_demux_summary || has_edx_concordance

# --- Derived demux analysis tibbles ---

# WDX on/off-target analysis
wdx_on_off <- tibble()
if (has_demux_summary && "barcode" %in% colnames(sample_metadata)) {
  expected_barcodes <- sample_metadata |>
    filter(!is.na(barcode)) |>
    distinct(run_path, barcode) |>
    mutate(run_id = basename(run_path))

  wdx_on_off <- demux_summary_all |>
    left_join(
      expected_barcodes |> dplyr::select(run_id, barcode),
      by = c("run_id", "predicted_barcode" = "barcode")
    ) |>
    mutate(
      category = case_when(
        predicted_barcode == "unclassified" ~ "Unclassified",
        !is.na(run_id) & predicted_barcode %in% expected_barcodes$barcode ~ "On-target",
        TRUE ~ "Off-target"
      )
    ) |>
    group_by(run_id, category) |>
    summarise(reads = sum(n_reads), .groups = "drop") |>
    group_by(run_id) |>
    mutate(
      total = sum(reads),
      pct = reads / total * 100
    ) |>
    ungroup() |>
    mutate(category = factor(category, levels = c("On-target", "Off-target", "Unclassified")))
}

# EDX on/off-target analysis
edx_on_off <- tibble()
if (has_edx_concordance && "edx" %in% colnames(sample_metadata)) {
  expected_edx_map <- sample_metadata |>
    filter(!is.na(edx)) |>
    dplyr::select(sample, expected_edx = edx)

  edx_on_off <- edx_concordance |>
    left_join(expected_edx_map, by = "sample") |>
    mutate(
      category = case_when(
        edx_adapter == "none" ~ "Undetected",
        edx_adapter == expected_edx ~ "On-target",
        TRUE ~ "Cross-talk"
      )
    ) |>
    group_by(sample, category) |>
    summarise(reads = sum(n_reads), .groups = "drop") |>
    group_by(sample) |>
    mutate(
      total = sum(reads),
      pct = reads / total * 100
    ) |>
    ungroup() |>
    mutate(
      sample = factor(sample, levels = sample_list),
      category = factor(category, levels = c("On-target", "Cross-talk", "Undetected"))
    )
}

# Demux funnel data: per-sample stage-by-stage read counts
demux_funnel <- tibble()
if (has_any_demux) {
  # Barcode-to-sample mapping and run mapping
  barcode_to_sample_map <- sample_metadata |>
    filter(!is.na(barcode)) |>
    mutate(run_id = basename(run_path)) |>
    dplyr::select(sample, barcode, run_id)

  funnel_rows <- list()
  for (s in sample_list) {
    stage_order <- 1
    sample_meta <- barcode_to_sample_map |> filter(sample == s)

    # Stage 1: Run Total (total reads in the run this sample came from)
    if (nrow(sample_meta) > 0 && has_demux_summary) {
      rid <- sample_meta$run_id[1]
      run_total <- demux_summary_all |> filter(run_id == rid) |> pull(n_reads) |> sum()
      funnel_rows <- c(funnel_rows, list(tibble(
        sample = s, stage = "Run Total", stage_order = stage_order, reads = run_total
      )))
      stage_order <- stage_order + 1

      # Stage 2: WDX On-target
      bc <- sample_meta$barcode[1]
      wdx_reads <- demux_summary_all |>
        filter(run_id == rid, predicted_barcode == bc) |>
        pull(n_reads) |>
        sum()
      funnel_rows <- c(funnel_rows, list(tibble(
        sample = s, stage = "WDX On-target", stage_order = stage_order, reads = wdx_reads
      )))
      stage_order <- stage_order + 1
    }

    # Stage 3: EDX On-target (if applicable)
    if (has_edx_concordance) {
      edx_on <- edx_on_off |> filter(sample == s, category == "On-target")
      if (nrow(edx_on) > 0) {
        funnel_rows <- c(funnel_rows, list(tibble(
          sample = s, stage = "EDX On-target", stage_order = stage_order, reads = edx_on$reads[1]
        )))
        stage_order <- stage_order + 1
      }
    }

    # Stage 4: Aligned
    aligned <- align_stats_all |> filter(sample == s, info == "aligned")
    if (nrow(aligned) > 0) {
      funnel_rows <- c(funnel_rows, list(tibble(
        sample = s, stage = "Aligned", stage_order = stage_order, reads = aligned$n_reads[1]
      )))
      stage_order <- stage_order + 1
    }

    # Stage 5: Classified
    classified <- align_stats_all |> filter(sample == s, info == "classified")
    if (nrow(classified) > 0) {
      funnel_rows <- c(funnel_rows, list(tibble(
        sample = s, stage = "Classified", stage_order = stage_order, reads = classified$n_reads[1]
      )))
    }
  }

  if (length(funnel_rows) > 0) {
    demux_funnel <- bind_rows(funnel_rows) |>
      mutate(
        sample = factor(sample, levels = sample_list),
        stage = fct_reorder(stage, stage_order)
      )
  }
}
```

```{r}
#| label: demux-not-available
#| eval: !expr (!has_any_demux)
#| results: asis

cat("*Demultiplexing was not enabled for this run.*\n")
```

```{r}
#| label: demux-wdx-on-off
#| eval: !expr has_demux_summary
#| results: asis

cat("\n\n### WDX On/Off-Target Analysis\n\n")
cat("Breakdown of WarpDemuX barcode assignments into on-target (expected barcodes), off-target (unexpected barcodes), and unclassified reads per run.\n\n")

# WDX on/off-target summary table
wdx_on_off_wide <- wdx_on_off |>
  dplyr::select(run_id, category, reads, pct) |>
  pivot_wider(
    id_cols = run_id,
    names_from = category,
    values_from = c(reads, pct),
    values_fill = list(reads = 0, pct = 0)
  )

# Ensure all expected columns exist
for (col in c("reads_On-target", "reads_Off-target", "reads_Unclassified",
              "pct_On-target", "pct_Off-target", "pct_Unclassified")) {
  if (!col %in% colnames(wdx_on_off_wide)) {
    wdx_on_off_wide[[col]] <- 0
  }
}

wdx_on_off_display <- wdx_on_off_wide |>
  mutate(Total = `reads_On-target` + `reads_Off-target` + `reads_Unclassified`) |>
  dplyr::select(
    Run = run_id,
    Total,
    `On-target` = `reads_On-target`,
    `On-target %` = `pct_On-target`,
    `Off-target` = `reads_Off-target`,
    `Off-target %` = `pct_Off-target`,
    Unclassified = `reads_Unclassified`,
    `Unclassified %` = `pct_Unclassified`
  )

tbl <- gt(wdx_on_off_display) |>
  fmt_number(columns = c(Total, `On-target`, `Off-target`, Unclassified), suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = c(`On-target %`, `Off-target %`, `Unclassified %`), decimals = 1, pattern = "{x}%")
cat(as_raw_html(tbl))
```

```{r}
#| label: fig-wdx-on-off-bar
#| eval: !expr has_demux_summary
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Stacked bar chart of WDX barcode classification per run. Green = on-target, orange = off-target, grey = unclassified."

ggplot(wdx_on_off, aes(x = run_id, y = reads, fill = category)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    position = position_stack(vjust = 0.5),
    size = 3, color = "white", fontface = "bold"
  ) +
  scale_fill_manual(values = c("On-target" = "#2d8a4e", "Off-target" = "#E69F00", "Unclassified" = "grey60")) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Number of Reads", fill = "Category") +
  theme_minimal_hgrid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: demux-wdx-detail
#| eval: !expr has_demux_summary
#| results: asis

cat("\n\n::: {.callout-note collapse=\"true\"}\n")
cat("#### Detailed WDX Barcode Table\n\n")

barcode_to_sample <- sample_metadata |>
  filter(!is.na(barcode)) |>
  dplyr::select(sample, barcode) |>
  deframe()

wdx_display <- demux_summary_all |>
  mutate(
    assigned_sample = barcode_to_sample[predicted_barcode] %||% NA_character_,
    assigned_sample = if_else(is.na(assigned_sample), "\u2014", assigned_sample)
  ) |>
  arrange(run_id, desc(n_reads)) |>
  dplyr::select(
    `Run` = run_id,
    `Barcode` = predicted_barcode,
    `Sample` = assigned_sample,
    `Reads` = n_reads
  )

tbl <- gt(wdx_display) |>
  fmt_number(columns = Reads, suffixing = TRUE, decimals = 1)
cat(as_raw_html(tbl))
cat("\n\n:::\n")
```

```{r}
#| label: demux-edx-on-off
#| eval: !expr has_edx_concordance
#| results: asis

cat("\n\n### EDX Adapter On/Off-Target Analysis\n\n")
cat("Breakdown of detected 3' adapter identity into on-target (matches expected EDX adapter), cross-talk (wrong adapter), and undetected reads per sample.\n\n")

# EDX on/off-target summary table
edx_on_off_wide <- edx_on_off |>
  dplyr::select(sample, category, reads, pct) |>
  pivot_wider(
    id_cols = sample,
    names_from = category,
    values_from = c(reads, pct),
    values_fill = list(reads = 0, pct = 0)
  )

for (col in c("reads_On-target", "reads_Cross-talk", "reads_Undetected",
              "pct_On-target", "pct_Cross-talk", "pct_Undetected")) {
  if (!col %in% colnames(edx_on_off_wide)) {
    edx_on_off_wide[[col]] <- 0
  }
}

edx_on_off_display <- edx_on_off_wide |>
  mutate(Total = `reads_On-target` + `reads_Cross-talk` + `reads_Undetected`) |>
  dplyr::select(
    Sample = sample,
    Total,
    `On-target` = `reads_On-target`,
    `On-target %` = `pct_On-target`,
    `Cross-talk` = `reads_Cross-talk`,
    `Cross-talk %` = `pct_Cross-talk`,
    Undetected = `reads_Undetected`,
    `Undetected %` = `pct_Undetected`
  )

tbl <- gt(edx_on_off_display) |>
  fmt_number(columns = c(Total, `On-target`, `Cross-talk`, Undetected), suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = c(`On-target %`, `Cross-talk %`, `Undetected %`), decimals = 1, pattern = "{x}%")
cat(as_raw_html(tbl))
```

```{r}
#| label: fig-edx-on-off-bar
#| eval: !expr has_edx_concordance
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Stacked bar chart of EDX adapter classification per sample. Green = on-target, red = cross-talk, grey = undetected."

ggplot(edx_on_off, aes(x = sample, y = reads, fill = category)) +
  geom_col(position = "stack", width = 0.7) +
  geom_text(
    aes(label = paste0(round(pct, 1), "%")),
    position = position_stack(vjust = 0.5),
    size = 3, color = "white", fontface = "bold"
  ) +
  scale_fill_manual(values = c("On-target" = "#2d8a4e", "Cross-talk" = "#c9372c", "Undetected" = "grey60")) +
  scale_y_continuous(labels = label_number(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.05))) +
  labs(x = NULL, y = "Number of Reads", fill = "Category") +
  theme_minimal_hgrid() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: demux-edx-detail
#| eval: !expr has_edx_concordance
#| results: asis

cat("\n\n::: {.callout-note collapse=\"true\"}\n")
cat("#### Detailed EDX Concordance Table\n\n")

expected_edx <- sample_metadata |>
  filter(!is.na(edx)) |>
  dplyr::select(sample, edx) |>
  deframe()

edx_display <- edx_concordance |>
  mutate(
    expected = expected_edx[sample],
    match = edx_adapter == expected
  )

tbl <- edx_display |>
  dplyr::select(
    Sample = sample,
    `Detected Adapter` = edx_adapter,
    `Expected` = expected,
    Reads = n_reads,
    `%` = pct
  ) |>
  gt() |>
  fmt_number(columns = Reads, suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = `%`, decimals = 1)

cat(as_raw_html(tbl))
cat("\n\n:::\n")
```

```{r}
#| label: demux-yield-funnel
#| eval: !expr has_any_demux
#| results: asis

cat("\n\n### Demux Yield Summary\n\n")
cat("Read attrition through each stage of the demultiplexing and processing pipeline.\n\n")
cat("\n\n::: {.panel-tabset}\n\n")

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  res <- knitr::knit_child(text = c(
    "```{r}",
    "#| echo: false",
    "#| fig-width: 10",
    "#| fig-height: 6",
    sprintf("plot_demux_funnel('%s', demux_funnel)", s),
    "```",
    ""
  ), envir = environment(), quiet = TRUE)
  cat(res, sep = "\n")
  cat("\n\n")
}

cat("\n\n:::\n\n")
```

```{r}
#| label: demux-efficiency-overview
#| eval: !expr has_any_demux
#| results: asis

cat("\n\n### Demux Efficiency Overview\n\n")
cat("Combined demultiplexing efficiency across all samples.\n\n")

# Build efficiency table from funnel data
efficiency_tbl <- demux_funnel |>
  dplyr::select(sample, stage, reads) |>
  pivot_wider(names_from = stage, values_from = reads)

# Determine which columns exist
has_run_total <- "Run Total" %in% colnames(efficiency_tbl)
has_wdx_col <- "WDX On-target" %in% colnames(efficiency_tbl)
has_edx_col <- "EDX On-target" %in% colnames(efficiency_tbl)
has_aligned_col <- "Aligned" %in% colnames(efficiency_tbl)
has_classified_col <- "Classified" %in% colnames(efficiency_tbl)

# Add run info
if ("barcode" %in% colnames(sample_metadata)) {
  run_map <- sample_metadata |>
    filter(!is.na(barcode)) |>
    mutate(run_id = basename(run_path)) |>
    dplyr::select(sample, run_id)
  efficiency_tbl <- efficiency_tbl |>
    left_join(run_map, by = "sample")
}

# Compute percentage columns
if (has_wdx_col && has_run_total) {
  efficiency_tbl <- efficiency_tbl |> mutate(`WDX %` = `WDX On-target` / `Run Total` * 100)
}
if (has_edx_col && has_wdx_col) {
  efficiency_tbl <- efficiency_tbl |> mutate(`EDX %` = `EDX On-target` / `WDX On-target` * 100)
}
if (has_aligned_col) {
  prev_col <- if (has_edx_col) "EDX On-target" else if (has_wdx_col) "WDX On-target" else "Run Total"
  if (prev_col %in% colnames(efficiency_tbl)) {
    efficiency_tbl <- efficiency_tbl |> mutate(`Align %` = Aligned / .data[[prev_col]] * 100)
  }
}
if (has_classified_col && has_aligned_col) {
  efficiency_tbl <- efficiency_tbl |> mutate(`Classify %` = Classified / Aligned * 100)
}
if (has_classified_col && has_run_total) {
  efficiency_tbl <- efficiency_tbl |> mutate(`Overall Yield %` = Classified / `Run Total` * 100)
}

# Select display columns in order
display_cols <- c("sample")
if ("run_id" %in% colnames(efficiency_tbl)) display_cols <- c(display_cols, "run_id")
if (has_run_total) display_cols <- c(display_cols, "Run Total")
if (has_wdx_col) display_cols <- c(display_cols, "WDX On-target", "WDX %")
if (has_edx_col) display_cols <- c(display_cols, "EDX On-target", "EDX %")
if (has_aligned_col) display_cols <- c(display_cols, "Aligned", "Align %")
if (has_classified_col) display_cols <- c(display_cols, "Classified", "Classify %")
if ("Overall Yield %" %in% colnames(efficiency_tbl)) display_cols <- c(display_cols, "Overall Yield %")

efficiency_display <- efficiency_tbl |>
  dplyr::select(any_of(display_cols))

# Rename for display
col_renames <- c(sample = "Sample")
if ("run_id" %in% colnames(efficiency_display)) col_renames <- c(col_renames, run_id = "Run")

# Build gt table
tbl <- gt(efficiency_display) |>
  cols_label(!!!col_renames)

# Format count columns
count_cols <- intersect(c("Run Total", "WDX On-target", "EDX On-target", "Aligned", "Classified"),
                        colnames(efficiency_display))
if (length(count_cols) > 0) {
  tbl <- tbl |> fmt_number(columns = all_of(count_cols), suffixing = TRUE, decimals = 1)
}

# Format percentage columns
pct_cols <- intersect(c("WDX %", "EDX %", "Align %", "Classify %", "Overall Yield %"),
                      colnames(efficiency_display))
if (length(pct_cols) > 0) {
  tbl <- tbl |> fmt_number(columns = all_of(pct_cols), decimals = 1, pattern = "{x}%")
}

# Hide EDX columns if no EDX
if (!has_edx_concordance) {
  edx_hide <- intersect(c("EDX On-target", "EDX %"), colnames(efficiency_display))
  if (length(edx_hide) > 0) {
    tbl <- tbl |> cols_hide(columns = all_of(edx_hide))
  }
}

cat(as_raw_html(tbl))
```

## Summary

```{r}
#| label: summary-table
#| tbl-cap: "Charging classification summary across all samples."

summary_tbl <- map_dfr(sample_list, function(s) {
  sample_prob <- charging_prob_all |> filter(sample == s)
  n_charged <- sum(sample_prob$charging_likelihood >= ml_threshold)
  n_uncharged <- sum(sample_prob$charging_likelihood < ml_threshold)
  n_total <- nrow(sample_prob)

  tibble(
    Sample = s,
    `Total Classified Reads` = n_total,
    `Charged Reads` = n_charged,
    `Uncharged Reads` = n_uncharged,
    `% Charged` = n_charged / n_total * 100
  )
})

gt(summary_tbl) |>
  fmt_number(columns = c(`Total Classified Reads`, `Charged Reads`, `Uncharged Reads`), suffixing = TRUE, decimals = 1) |>
  fmt_number(columns = `% Charged`, decimals = 1)
```

**ML threshold used:** `r ml_threshold`

{{< include _custom.qmd >}}

## Appendix

### Reference Sequence Similarity

This section shows pairwise sequence similarity across all sequences in the reference FASTA file. High similarity between sequences may indicate potential cross-mapping issues during alignment.

```{r}
#| label: reference-similarity-check
#| include: false

# Check if similarity matrix is available
has_similarity_matrix <- !is.null(similarity_matrix)
```

```{r}
#| label: fig-similarity-heatmap
#| fig-width: 12
#| fig-height: 10
#| fig-cap: "Pairwise sequence identity heatmap. Amino acid annotations shown on axes. High similarity (>90%) shown in orange/red."
#| eval: !expr has_similarity_matrix

# Parse tRNA names to extract amino acid and anticodon
parse_trna_name <- function(name) {
  parts <- str_split(name, "-")[[1]]
  if (length(parts) >= 3 && parts[1] == "tRNA") {
    aa <- parts[2]
    anticodon <- parts[3]
  } else if (length(parts) >= 2) {
    aa <- parts[1]
    anticodon <- if (length(parts) >= 2) parts[2] else "unknown"
  } else {
    aa <- "unknown"
    anticodon <- "unknown"
  }
  list(aa = aa, anticodon = anticodon)
}

# Parse all sequence names
parsed <- lapply(rownames(similarity_matrix), parse_trna_name)
aa_labels <- sapply(parsed, function(x) x$aa)

# Create color palette for amino acids
unique_aa <- unique(aa_labels)
n_aa <- length(unique_aa)

if (n_aa <= 20) {
  aa_colors <- setNames(rainbow(n_aa, s = 0.7, v = 0.8), unique_aa)
} else {
  aa_colors <- setNames(
    colorRampPalette(c("#E41A1C", "#377EB8", "#4DAF4A", "#984EA3",
                       "#FF7F00", "#FFFF33", "#A65628", "#F781BF"))(n_aa),
    unique_aa
  )
}

# Color scale for similarity values
col_fun <- colorRamp2(c(50, 75, 90, 100), c("white", "lightyellow", "orange", "red"))

# Create annotations
row_anno <- rowAnnotation(
  AA = aa_labels,
  col = list(AA = aa_colors),
  show_annotation_name = TRUE,
  annotation_name_side = "top"
)

col_anno <- HeatmapAnnotation(
  AA = aa_labels,
  col = list(AA = aa_colors),
  show_annotation_name = FALSE
)

# Draw heatmap
ht <- Heatmap(
  similarity_matrix,
  name = "% Identity",
  col = col_fun,
  left_annotation = row_anno,
  top_annotation = col_anno,
  show_row_names = nrow(similarity_matrix) <= 50,
  show_column_names = nrow(similarity_matrix) <= 50,
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 6),
  clustering_distance_rows = "euclidean",
  clustering_distance_columns = "euclidean",
  clustering_method_rows = "complete",
  clustering_method_columns = "complete",
  row_title = "Reference Sequences",
  column_title = "Reference Sequence Similarity"
)

draw(ht)
```

```{r}
#| label: no-similarity-matrix
#| eval: !expr (!has_similarity_matrix)
#| results: asis

cat("*Reference similarity matrix not available.*\n")
```

#### Highly Similar Sequence Pairs

Sequence pairs with >90% identity may be at risk for cross-mapping:

```{r}
#| label: high-similarity-pairs
#| eval: !expr has_similarity_matrix

high_sim_threshold <- 90

# Find pairs with high similarity (excluding diagonal).
# Indexed rather than looped: a nested loop over the matrix is O(n^2)
# interpreted iterations and does not finish on a large reference.
hits <- which(
  similarity_matrix >= high_sim_threshold & upper.tri(similarity_matrix),
  arr.ind = TRUE
)

pairs <- data.frame(
  Seq1 = rownames(similarity_matrix)[hits[, "row"]],
  Seq2 = colnames(similarity_matrix)[hits[, "col"]],
  Percent_Identity = round(similarity_matrix[hits], 2)
)

# Sort by similarity (descending)
pairs <- pairs |> arrange(desc(Percent_Identity))

if (nrow(pairs) > 0) {
  gt(head(pairs, 50)) |>
    cols_label(
      Seq1 = "Sequence 1",
      Seq2 = "Sequence 2",
      Percent_Identity = "% Identity"
    ) |>
    fmt_number(columns = Percent_Identity, decimals = 1)
} else {
  cat(paste0("No sequence pairs found with >= ", high_sim_threshold, "% identity.\n"))
}
```

#### Similarity Distribution

```{r}
#| label: fig-similarity-histogram
#| fig-width: 8
#| fig-height: 4
#| fig-cap: "Distribution of pairwise sequence similarities. Red dashed line indicates 90% threshold."
#| eval: !expr has_similarity_matrix

# Get upper triangle values (excluding diagonal)
upper_tri <- similarity_matrix[upper.tri(similarity_matrix)]

tibble(similarity = upper_tri) |>
  ggplot(aes(x = similarity)) +
  geom_histogram(binwidth = 2, fill = "steelblue", color = "white", alpha = 0.8) +
  geom_vline(xintercept = 90, linetype = "dashed", color = "red", linewidth = 1) +
  scale_y_continuous(labels = comma) +
  labs(x = "Percent Identity", y = "Number of Pairs") +
  theme_minimal_hgrid()
```

```{r}
#| label: similarity-summary-stats
#| eval: !expr has_similarity_matrix

upper_tri <- similarity_matrix[upper.tri(similarity_matrix)]

tibble(
  Statistic = c("Minimum", "Median", "Mean", "Maximum", "Pairs >= 90%", "Pairs >= 95%", "Total pairs"),
  Value = c(
    sprintf("%.1f%%", min(upper_tri)),
    sprintf("%.1f%%", median(upper_tri)),
    sprintf("%.1f%%", mean(upper_tri)),
    sprintf("%.1f%%", max(upper_tri)),
    as.character(sum(upper_tri >= 90)),
    as.character(sum(upper_tri >= 95)),
    as.character(length(upper_tri))
  )
) |>
  gt()
```

### Sample Details

```{r}
#| label: sample-details

if ("barcode" %in% colnames(sample_metadata)) {
  cols <- c(Sample = "sample", `Run Path` = "run_path",
            Barcode = "barcode", `Barcode Kit` = "barcode_kit")
  if ("edx" %in% colnames(sample_metadata) && any(!is.na(sample_metadata$edx))) {
    cols <- c(cols, `EDX Adapter` = "edx")
  }
  sample_metadata |>
    dplyr::select(!!!cols) |>
    gt()
} else {
  sample_metadata |>
    dplyr::select(Sample = sample, `Run Path` = run_path) |>
    gt()
}
```

### Pipeline Manifest

```{r}
#| label: manifest-check
#| include: false

manifest_path <- fs::path(output_dir, "manifest.json")
has_manifest <- fs::file_exists(manifest_path)
```

```{r}
#| label: manifest-table
#| eval: !expr has_manifest

manifest <- yaml::read_yaml(manifest_path)

# Format git commit as short hash
short_commit <- if (!is.null(manifest$pipeline$git_commit)) {
  substr(manifest$pipeline$git_commit, 1, 7)
} else "N/A"

# Format dirty flag
dirty_label <- if (isTRUE(manifest$pipeline$git_dirty)) "Yes" else "No"

# Format timestamps
fmt_time <- function(ts) {
  if (is.null(ts)) return("N/A")
  format(as.POSIXct(ts, format = "%Y-%m-%dT%H:%M:%S", tz = "UTC"), "%Y-%m-%d %H:%M:%S UTC")
}

# Format 3' adapter (may be list or string)
three_prime_adapter <- if (is.list(manifest$config$adapters$three_prime)) {
  paste(sapply(manifest$config$adapters$three_prime,
    function(x) paste0(x$name, ": ", x$seq)), collapse = "; ")
} else {
  manifest$config$adapters$three_prime %||% "N/A"
}

# Determine reference FASTA display based on reference mode
ref_mode <- manifest$config$reference$mode %||% "validate"
if (ref_mode == "build") {
  ref_fasta_label <- "Reference Source"
  ref_fasta_value <- manifest$config$reference$raw_fasta %||% "N/A"
} else {
  ref_fasta_label <- "Reference FASTA"
  ref_fasta_value <- manifest$config$fasta %||% "N/A"
}

# Build a single combined table
manifest_tbl <- tribble(
  ~Group, ~Field, ~Value,
  # Execution
  "Execution", "Status",   manifest$execution$status %||% "N/A",
  "Execution", "Started",  fmt_time(manifest$execution$timestamp_start),
  "Execution", "Finished", fmt_time(manifest$execution$timestamp_end),
  "Execution", "Host",     manifest$execution$hostname %||% "N/A",
  "Execution", "User",     manifest$execution$user %||% "N/A",
  # Pipeline version
  "Pipeline", "Git Commit",          short_commit,
  "Pipeline", "Branch",              manifest$pipeline$git_branch %||% "N/A",
  "Pipeline", "Tag",                 manifest$pipeline$git_tag %||% "N/A",
  "Pipeline", "Uncommitted Changes", dirty_label,
  # Configuration
  "Configuration", ref_fasta_label,          ref_fasta_value,
  "Configuration", "Reference Mode",         ref_mode,
  "Configuration", "Base Calling Model",    manifest$config$dorado_model %||% "N/A",
  "Configuration", "Dorado Options",        trimws(manifest$config$opts$dorado %||% "N/A"),
  "Configuration", "BWA Options",           trimws(manifest$config$opts$bwa %||% "N/A"),
  "Configuration", "BAM Filter",            trimws(manifest$config$opts$bam_filter %||% "N/A"),
  "Configuration", "5' Adapter",            manifest$config$adapters$five_prime %||% "N/A",
  "Configuration", "3' Adapter",            three_prime_adapter,
  "Configuration", "Charging ML Threshold", as.character(ml_threshold),
)

# Append WarpDemuX row if enabled
if (isTRUE(manifest$config$warpdemux$enabled)) {
  manifest_tbl <- bind_rows(manifest_tbl, tibble(
    Group = "Configuration",
    Field = "WarpDemuX Barcode Kit",
    Value = manifest$config$warpdemux$barcode_kit %||% "N/A"
  ))
}

# Append tool versions
tool_rows <- tibble(
  Group = "Tools",
  Field = names(manifest$tools),
  Value = unlist(manifest$tools)
)
manifest_tbl <- bind_rows(manifest_tbl, tool_rows)

# Render table
manifest_tbl |>
  gt(groupname_col = "Group") |>
  cols_label(Field = "", Value = "") |>
  tab_style(
    style = cell_text(font = "monospace", size = "small"),
    locations = cells_body(columns = Value,
      rows = Field %in% c("Git Commit", "Dorado Options", "BWA Options",
                           "BAM Filter", "5' Adapter", "3' Adapter"))
  ) |>
  tab_style(
    style = cell_text(weight = "bold", color = if (manifest$execution$status == "success") "#2d8a4e" else "#c9372c"),
    locations = cells_body(columns = Value, rows = Field == "Status")
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) |>
  tab_footnote(
    footnote = manifest$pipeline$git_commit %||% "",
    locations = cells_body(columns = Value, rows = Field == "Git Commit")
  ) |>
  (\(tbl) {
    if (ref_mode == "build") {
      tbl |> tab_footnote(
        footnote = "Adapted reference built from raw tRNA sequences with pipeline adapters prepended/appended",
        locations = cells_body(columns = Value, rows = Field == "Reference Mode")
      )
    } else tbl
  })() |>
  tab_footnote(
    footnote = "Requires reads with 24bp 5' adapter, 23bp 3' adapter, positive strand",
    locations = cells_body(columns = Value, rows = Field == "BAM Filter")
  ) |>
  tab_footnote(
    footnote = paste0("Reads with ML >= ", ml_threshold, " classified as charged"),
    locations = cells_body(columns = Value, rows = Field == "Charging ML Threshold")
  ) |>
  tab_options(
    row_group.padding = px(4),
    data_row.padding = px(3),
    footnotes.font.size = "small"
  )
```

```{r}
#| label: no-manifest
#| eval: !expr (!has_manifest)
#| results: asis

cat("*Pipeline manifest not available. Manifest is generated on pipeline completion.*\n")
```

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info

sessioninfo::session_info()
```

:::
