---
title: "aa-tRNA-seq Combined QC Report"
author: "Pipeline"
date: today
format:
  html:
    code-fold: true
    code-summary: "Show code"
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  warning: false
  message: false
params:
  config_file: "config/config.yml"
  ml_threshold: 200
---

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(gt)
library(paletteer)
library(fs)
library(patchwork)
library(scales)
library(yaml)
library(cowplot)

# Read pipeline config
config <- read_yaml(params$config_file)
output_dir <- fs::path_abs(config$output_directory, start = dirname(params$config_file))
ml_threshold <- params$ml_threshold

# Parse samples from config
# Supports both YAML format (with runs) and TSV format
if (!is.null(config$samples))
{
  samples_path <- fs::path_abs(config$samples, start = dirname(params$config_file))
  if (fs::path_ext(samples_path) %in% c("yml", "yaml")) {
    samples_config <- read_yaml(samples_path)
    sample_list <- samples_config$runs |>
      map(~ names(.x$samples)) |>
      unlist()
  } else {
    # TSV format: first column is sample name
    sample_list <- read_tsv(samples_path, col_names = FALSE, show_col_types = FALSE)$X1
  }
} else {
 stop("No samples defined in config file")
}

# Extract isodecoder from gtrnadb-format tRNA names
# Pattern: Genus_species_tRNA-Aaa-NNN-X-Y → Aaa-NNN (handles iMet with 4 chars)
# Returns list with isodecoder name and whether it matched the tRNA pattern
extract_isodecoder <- function(trna_name) {
  isodecoder <- str_extract(trna_name, "(?<=tRNA-)[A-Za-z]{3,4}-[ACGT]{3}")
  if_else(is.na(isodecoder), trna_name, isodecoder)
}

# Check if name matches standard tRNA isodecoder pattern
is_standard_trna <- function(trna_name) {
 str_detect(trna_name, "tRNA-[A-Za-z]{3,4}-[ACGT]{3}")
}
```

This report summarizes QC metrics for **`r length(sample_list)` samples** from the aa-tRNA-seq pipeline.

**Samples:** `r paste(sample_list, collapse = ", ")`

```{r}
#| label: load-all-data
#| include: false

# Load all data upfront
align_stats_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".align_stats.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

charging_prob_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".charging_prob.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

charging_cpm_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".charging.cpm.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})

bcerror_all <- map_dfr(sample_list, function(s) {
  tables_dir <- fs::path(output_dir, "summary", "tables", s)
  read_tsv(fs::path(tables_dir, paste0(s, ".bcerror.tsv.gz")), show_col_types = FALSE) |>
    mutate(sample = s)
})
```

## Alignment Summary

### Read Progression Through Pipeline

::: {.panel-tabset}

```{r}
#| label: alignment-summary-tabs
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_align <- align_stats_all |>
    filter(sample == s) |>
    dplyr::select(
      Stage = info,
      `Total Reads` = n_reads,
      `Mapped Reads` = mapped_reads,
      `% Mapped` = pct_mapped,
      `Mean Length` = mean_length,
      `Median Length` = median_length,
      `Mean Quality` = mean_base_quality,
      `Mean MAPQ` = mean_MAPQ
    ) |>
    mutate(
      Stage = factor(Stage, levels = c("unmapped", "aligned", "classified")),
      Stage = recode(Stage,
        "unmapped" = "Rebasecalled",
        "aligned" = "Aligned",
        "classified" = "Classified"
      )
    )

  tbl <- gt(tbl_align) |>
    fmt_number(columns = c(`Mean Length`, `Mean Quality`, `Mean MAPQ`), decimals = 1) |>
    fmt_number(columns = `% Mapped`, decimals = 1)

  print(htmltools::tagList(tbl))
  cat("\n\n")
}
```

:::

### Alignment Funnel

```{r}
#| label: alignment-funnel
#| fig-width: 10
#| fig-height: 4
#| fig-cap: "Read counts at each processing stage showing retention through the pipeline."

tbl_funnel <- align_stats_all |>
  dplyr::select(sample, info, n_reads) |>
  mutate(
    stage = factor(info, levels = c("unmapped", "aligned", "classified")),
    stage = recode(stage,
      "unmapped" = "Rebasecalled",
      "aligned" = "Aligned",
      "classified" = "Classified"
    )
  )

ggplot(tbl_funnel, aes(x = stage, y = n_reads, fill = stage)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = comma(n_reads)), vjust = -0.5, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Processing Stage", y = "Number of Reads") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 45, hjust = 1))
```

## Charging Classification

### Charging Likelihood Distribution

The Remora model assigns a charging likelihood score (0-255) to each read. Reads with scores >= `r ml_threshold` are classified as charged.

```{r}
#| label: charging-density
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Distribution of Remora charging likelihood scores. Dashed line indicates classification threshold."

tbl_density <- charging_prob_all |>
  mutate(
    charging_status = if_else(charging_likelihood >= ml_threshold, "Charged", "Uncharged")
  )

ggplot(tbl_density, aes(x = charging_likelihood, fill = charging_status)) +
  geom_histogram(binwidth = 5, alpha = 0.8, position = "identity") +
  geom_vline(xintercept = ml_threshold, linetype = "dashed", linewidth = 1, color = "grey30") +
  scale_fill_manual(values = c("Charged" = "#E69F00", "Uncharged" = "#56B4E9")) +
  scale_y_continuous(labels = comma) +
  labs(x = "Charging Likelihood (ML tag, 0-255)", y = "Number of Reads", fill = "Classification") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid() +
  theme(legend.position = "top")
```

### Charging Summary

::: {.panel-tabset}

```{r}
#| label: charging-summary-tabs
#| results: asis

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  sample_prob <- charging_prob_all |> filter(sample == s)
  n_charged <- sum(sample_prob$charging_likelihood >= ml_threshold)
  n_uncharged <- sum(sample_prob$charging_likelihood < ml_threshold)
  n_total <- nrow(sample_prob)

  tbl_charging_summary <- tibble(
    Status = c("Charged", "Uncharged", "Total"),
    `Read Count` = c(n_charged, n_uncharged, n_total),
    Percentage = c(
      n_charged / n_total * 100,
      n_uncharged / n_total * 100,
      100
    )
  )

  tbl <- gt(tbl_charging_summary) |>
    fmt_number(columns = Percentage, decimals = 1) |>
    fmt_number(columns = `Read Count`, use_seps = TRUE, decimals = 0)

  print(htmltools::tagList(tbl))
  cat("\n\n")
}
```

:::

### Per-tRNA Abundance

::: {.panel-tabset}

```{r}
#| label: trna-abundance-tabs
#| results: asis
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Abundance (CPM) by tRNA isodecoder. Isodecoders with fewer than 5 total reads are excluded. Red bars indicate non-standard tRNA names."

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_abundance <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(
      is_standard = is_standard_trna(tRNA),
      isodecoder = extract_isodecoder(tRNA)
    ) |>
    group_by(isodecoder) |>
    summarise(
      total_counts = sum(counts_charged) + sum(counts_uncharged),
      total_cpm = sum(cpm_charged) + sum(cpm_uncharged),
      is_standard = first(is_standard),
      .groups = "drop"
    ) |>
    filter(total_counts >= 5) |>
    mutate(isodecoder = fct_reorder(isodecoder, total_cpm))

  if (nrow(tbl_abundance) > 0) {
    p <- ggplot(tbl_abundance, aes(x = isodecoder, y = total_cpm, fill = is_standard)) +
      geom_col() +
      scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E41A1C"), guide = "none") +
      scale_y_continuous(labels = comma) +
      labs(x = NULL, y = "CPM") +
      theme_minimal_hgrid() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.8)))

    print(p)
  } else {
    cat("No tRNAs with >= 5 reads\n")
  }
  cat("\n\n")
}
```

:::

### Per-tRNA Charging

::: {.panel-tabset}

```{r}
#| label: trna-charging-tabs
#| results: asis
#| fig-width: 10
#| fig-height: 5
#| fig-cap: "Charging percentage by tRNA isodecoder. Isodecoders with fewer than 5 total reads are excluded. Red bars indicate non-standard tRNA names."

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_charging <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(
      is_standard = is_standard_trna(tRNA),
      isodecoder = extract_isodecoder(tRNA)
    ) |>
    group_by(isodecoder) |>
    summarise(
      counts_charged = sum(counts_charged),
      counts_uncharged = sum(counts_uncharged),
      is_standard = first(is_standard),
      .groups = "drop"
    ) |>
    mutate(
      total_counts = counts_charged + counts_uncharged,
      pct_charged = counts_charged / total_counts * 100
    ) |>
    filter(total_counts >= 5) |>
    mutate(isodecoder = fct_reorder(isodecoder, pct_charged))

  if (nrow(tbl_charging) > 0) {
    p <- ggplot(tbl_charging, aes(x = isodecoder, y = pct_charged, fill = is_standard)) +
      geom_col() +
      scale_fill_manual(values = c("TRUE" = "grey60", "FALSE" = "#E41A1C"), guide = "none") +
      scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, 25)) +
      labs(x = NULL, y = "% Charged") +
      theme_minimal_hgrid() +
      theme(axis.text.x = element_text(angle = 45, hjust = 1, size = rel(0.8)))

    print(p)
  } else {
    cat("No tRNAs with >= 5 reads\n")
  }
  cat("\n\n")
}
```

:::

### Top tRNAs by Read Count

::: {.panel-tabset}

```{r}
#| label: top-trnas-tabs
#| results: asis
#| tbl-cap: "Top 20 tRNA isodecoders ranked by total read count."

for (s in sample_list) {
  cat("\n\n####", s, "\n\n")

  tbl_top_trnas <- charging_cpm_all |>
    filter(sample == s) |>
    mutate(isodecoder = extract_isodecoder(tRNA)) |>
    group_by(isodecoder) |>
    summarise(
      counts_charged = sum(counts_charged),
      counts_uncharged = sum(counts_uncharged),
      cpm_charged = sum(cpm_charged),
      cpm_uncharged = sum(cpm_uncharged),
      .groups = "drop"
    ) |>
    mutate(
      total_counts = counts_charged + counts_uncharged,
      total_cpm = cpm_charged + cpm_uncharged,
      pct_charged = counts_charged / total_counts * 100
    ) |>
    arrange(desc(total_counts)) |>
    head(20) |>
    dplyr::select(
      Isodecoder = isodecoder,
      `Charged` = counts_charged,
      `Uncharged` = counts_uncharged,
      `Total` = total_counts,
      `% Charged` = pct_charged,
      `CPM` = total_cpm
    )

  tbl <- gt(tbl_top_trnas) |>
    fmt_number(columns = c(`Charged`, `Uncharged`, `Total`), use_seps = TRUE, decimals = 0) |>
    fmt_number(columns = `% Charged`, decimals = 1) |>
    fmt_number(columns = CPM, decimals = 0)

  print(htmltools::tagList(tbl))
  cat("\n\n")
}
```

:::

## Base Calling Quality

### Error Frequency by Position

```{r}
#| label: error-frequency
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Mean error frequency across all tRNAs by reference position."

tbl_error <- bcerror_all |>
  group_by(sample, Position) |>
  summarise(
    mean_mismatch = mean(MismatchFreq, na.rm = TRUE),
    mean_insertion = mean(InsertionFreq, na.rm = TRUE),
    mean_deletion = mean(DeletionFreq, na.rm = TRUE),
    .groups = "drop"
  )

tbl_error_long <- tbl_error |>
  dplyr::select(sample, Position, Mismatch = mean_mismatch, Insertion = mean_insertion, Deletion = mean_deletion) |>
  pivot_longer(cols = c(Mismatch, Insertion, Deletion), names_to = "Error Type", values_to = "Frequency")

ggplot(tbl_error_long, aes(x = Position, y = Frequency, color = `Error Type`)) +
  geom_line(linewidth = 0.8) +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent_format(scale = 100)) +
  labs(x = "Position in Reference", y = "Error Frequency") +
  facet_wrap(~sample) +
  theme_minimal_hgrid() +
  theme(legend.position = "top")
```

### Coverage by Position

```{r}
#| label: coverage-position
#| fig-width: 10
#| fig-height: 6
#| fig-cap: "Mean read coverage across all tRNAs by reference position."

tbl_coverage <- bcerror_all |>
  group_by(sample, Position) |>
  summarise(
    mean_coverage = mean(Spanning_Reads, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(tbl_coverage, aes(x = Position, y = mean_coverage)) +
  geom_area(fill = "#56B4E9", alpha = 0.6) +
  geom_line(color = "#0072B2", linewidth = 0.5) +
  scale_y_continuous(labels = comma) +
  labs(x = "Position in Reference", y = "Mean Coverage (Reads)") +
  facet_wrap(~sample, scales = "free_y") +
  theme_minimal_hgrid()
```

## Summary

```{r}
#| label: summary-table
#| tbl-cap: "Charging classification summary across all samples."

summary_tbl <- map_dfr(sample_list, function(s) {
  sample_prob <- charging_prob_all |> filter(sample == s)
  n_charged <- sum(sample_prob$charging_likelihood >= ml_threshold)
  n_uncharged <- sum(sample_prob$charging_likelihood < ml_threshold)
  n_total <- nrow(sample_prob)

  tibble(
    Sample = s,
    `Total Classified Reads` = n_total,
    `Charged Reads` = n_charged,
    `Uncharged Reads` = n_uncharged,
    `% Charged` = n_charged / n_total * 100
  )
})

gt(summary_tbl) |>
  fmt_number(columns = c(`Total Classified Reads`, `Charged Reads`, `Uncharged Reads`), use_seps = TRUE, decimals = 0) |>
  fmt_number(columns = `% Charged`, decimals = 1)
```

**ML threshold used:** `r ml_threshold`

{{< include _custom.qmd >}}

::: {.callout-note collapse="true"}
## Session Info

```{r}
#| label: session-info

sessioninfo::session_info()
```

:::
